### 2024-03-21 paper/assets/features_table.tex
Initial paper draft

\begin{table}
\centering
\caption{Features used in modeling and analysis}
\label{tab:features}
\begin{tabular}{llll}
\toprule
 & feature & entity & kind \\
\midrule
1 & disease\_\_clinical\_\_phase\_max\_\_reached & disease & temporal \\
2 & target\_\_clinical\_\_phase\_max\_\_reached & target & temporal \\
3 & target\_\_genetic\_constraint & target & static \\
4 & target\_\_mouse\_ko\_score & target & static \\
5 & target\_\_tissue\_distribution & target & static \\
6 & target\_\_tissue\_specificity & target & static \\
7 & target\_disease\_\_affected\_pathway\_\_cancer\_biomarkers & target\_disease & temporal \\
8 & target\_disease\_\_affected\_pathway\_\_crispr & target\_disease & temporal \\
9 & target\_disease\_\_affected\_pathway\_\_crispr\_screen & target\_disease & temporal \\
10 & target\_disease\_\_affected\_pathway\_\_progeny & target\_disease & temporal \\
11 & target\_disease\_\_affected\_pathway\_\_reactome & target\_disease & temporal \\
12 & target\_disease\_\_affected\_pathway\_\_slapenrich & target\_disease & temporal \\
13 & target\_disease\_\_affected\_pathway\_\_sysbio & target\_disease & temporal \\
14 & target\_disease\_\_animal\_model\_\_impc & target\_disease & temporal \\
15 & target\_disease\_\_genetic\_association\_\_clingen & target\_disease & static \\
16 & target\_disease\_\_genetic\_association\_\_curated & target\_disease & static \\
17 & target\_disease\_\_genetic\_association\_\_eva & target\_disease & static \\
18 & target\_disease\_\_genetic\_association\_\_gene2phenotype & target\_disease & static \\
19 & target\_disease\_\_genetic\_association\_\_gene\_burden & target\_disease & static \\
20 & target\_disease\_\_genetic\_association\_\_genomics\_england & target\_disease & static \\
21 & target\_disease\_\_genetic\_association\_\_omim & target\_disease & static \\
22 & target\_disease\_\_genetic\_association\_\_orphanet & target\_disease & static \\
23 & target\_disease\_\_genetic\_association\_\_ot\_genetics\_portal & target\_disease & static \\
24 & target\_disease\_\_genetic\_association\_\_uniprot\_literature & target\_disease & static \\
25 & target\_disease\_\_genetic\_association\_\_uniprot\_variants & target\_disease & static \\
26 & target\_disease\_\_known\_drug\_\_chembl & target\_disease & temporal \\
27 & target\_disease\_\_literature\_\_europepmc & target\_disease & temporal \\
28 & target\_disease\_\_outcome\_\_advanced & target\_disease & temporal \\
29 & target\_disease\_\_rna\_expression\_\_expression\_atlas & target\_disease & temporal \\
30 & target\_disease\_\_somatic\_mutation\_\_cancer\_gene\_census & target\_disease & temporal \\
31 & target\_disease\_\_somatic\_mutation\_\_eva\_somatic & target\_disease & temporal \\
32 & target\_disease\_\_somatic\_mutation\_\_intogen & target\_disease & temporal \\
33 & target\_disease\_\_time\_\_transition & target\_disease & temporal \\
\bottomrule
\end{tabular}
\end{table}

### 2024-03-21 paper/paper.tex
Initial paper draft

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\usepackage[table, x11names, svgnames]{xcolor}
\usepackage{subfig}
\usepackage[section]{placeins}
\usepackage{setspace}

\setstretch{1.2} % line spacing

\graphicspath{{assets/}}

%% Header
\pagestyle{fancy}
\thispagestyle{empty}
\rhead{ \textit{ }} 
\fancyhead[LO]{}

% \fancyhead[RE]{Firstauthor and Secondauthor} % Firstauthor et al. if more than 2 - must use \documentclass[twoside]{article}

%% Title
\title{Clinical advancement forecasting
%% Cite as
\thanks{\textit{\underline{Citation}}: 
\textbf{TBD}} 
}

%% Authors
\author{
  Authors TBD \\
  Related Sciences \\
%   %% examples of more authors
%    \And
%   Author3 \\
%   Affiliation \\
%   Univ \\
%   City\\
%   \texttt{email@email} \\
}

%% Body
\setlength{\parindent}{20pt}

\begin{document}
\maketitle

\begin{abstract}
This study examines the extent to which the outcomes of clinical trials can be predicted based on longitudinal properties of drug targets and diseases alone. We find that this is possible by comparing the historical performance of model-based target-disease pair priortiziation methods to common baselines. Our primary objective is to demonstrate that statistical learning can effectively optimize such methods with no loss of interpretability. For example, non-negative linear models can produce simple weighting schemes across various types of human, animal and cell model evidence (for targets, diseases and pairings of the two) to identify target-disease pairs that advance beyond phase 2 trials with an average relative risk that is 2x higher than Open Targets composite scores. Other key characteristics of this study include: 1) a comprehensive longitudinal treatment of evidence as well as how it relates to leakage and reverse causality in biomedical research, 2) trial and/or drug details are not used in order to enable ranking for undeveloped targets/diseases, 3) analysis of the space of currently undeveloped, tractable targets with the highest likelihood of clinical success, 3) no data is used outside of Open Targets to ease reproduction and/or deployment, and 4) our method requires no expert knowledge and can easily support the inclusion of more lines of evidence over time, making it easy to operationalize.
% Note: OT multiplier (2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10
\end{abstract}

\section{Introduction}

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:36963162}. This information has been used to devise many target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets/diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. To our knolwedge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates. This is not a focus of this work as we aim, instead, to establish a framework for identifying targets/diseases with the very highest possible likelihood of success first. This is accomplished by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in a simple statistical modeling framework. This can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret. We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence from many sources, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.

A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine. This is vital because any method that optimizes for likely future clinical success based on historical clinical success may easily be biased by the non-random nature with which evidence is absent otherwise. We use only "temporalized" evidence, i.e. evidence for which timing of its emergence can be determined, and outcomes when training and evaluating our methods before ultimately applying them to present-day evidence with no restrictions on timing. We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.

Addressing such problems is common, but not ubuquitous, in studies that attempt to predict clincal trial outcomes using temporalized predictors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}. The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets/diseases, trial sponsor track records, eventual patient enrollment, etc. constitute clear information leaks otherwise. This is discussed more in \cite{PMID:37483175} which notes several studies that do not account for this problem, and presents both "quasi-prospective" as well as true "prospective" results. The difference between these two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until years later. Nomenclature for these formulations is conflicting though, where this definition of a "quasi-prospective" design is deemed entirely prospective in some cases, e.g. \cite{PMID:37225853}. We will refer to our design in this study as quasi-prospective since this definition is the best fit.

The prior works discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with clinical trial success and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how the former methods are {\bf not} directly optimized for those outcomes while the latter methods are. In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are undeveloped (i.e. have never been in clinical trials), have a tractable target and are likely to see success rates matching or exceeding those calibrated benchmarks.

\section{Results}
\label{sec:results}

In order to model clinical advancement for target-disease (TD) pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with any one TD pair as indicated by the presence of a later-stage trial. All results to follow consider only advancement beyond phase 2 due to limitations described in Section~\ref{sec:discussion}. This binary outcome is then predicted based on a list of features shown in Supplementary~Table~\ref{tab:features}. Information for each of these features is only used when it was published before the year \textbf{prior} to the first phase 2 trial observed, with an exception for genetic evidence discussed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where this first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as lower as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.

\subsection{Features}

The predictors used consist of 27 target-disease pair features, 5 target-specific features and 1 disease-specific feature. These are listed in Supplementary~Table~\ref{tab:features}. The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods that do not favor well-studied or drugged targets/diseases. Examples of this include target-specific tissue expression specificity scores computed from Human Protein Atlas \cite{PMID:25613900} and LOEUF \cite{PMID:32461654} scores from gnomAD. Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OT23.12release} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features.

\subsection{Models}

We train a variety of models including constrained and unconstrained linear and tree models. The constrained variants of these models force effects of all features to increase monotonically. This is possible with no feature transformations because all scores in Open Targets are constructed such that higher scores are presumed to be advantageous. We also apply these models to our evaluation dataset using several feature ablations in order to assess the value of groups of related features.

In order to compare these models to an Open Targets composite score, we use an equally weighted sum of all scores except for those assigned lower weights in \cite{OTweights}. Scores from these sources are multiplied by the corresponding weight before being summed.

\subsection{Performance}
\label{sec:results_performance}

Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study -- a constrained, linear L2-regularized model referred to from here on as "RDG" -- ranks TD pairs by comparison to a composite score from Open Targets ("OTS"). The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:

\begin{equation}
  \frac{P(advancement | rank >= N)}{P(advancement | rank < N)}.
\end{equation}

The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it also uses time since the phase 2 transition as a predictive factor as well as all others. We observe that the use of this information almost universally improves standard performance metrics like ROC (receiver operating characteristic) and AP (average precision), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed. In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support. This is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion}). As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available. They are, however, a useful performance ceiling towards which future work might build towards.

A subset of therapeutic areas was used for this analysis and this is reflected in both the mean estimates and distributions of Figure~\ref{fig:performance_across_ta}. A full list of these can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta}, which also shows the the RR estimates used in any summary statistics across therapeutic areas. These were selected based on criteria described in Section~\ref{sec:methods}. Lastly, Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} shows RR distributions by model with significances of their differences.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \subfloat[\centering]{\includegraphics[width=.58\textwidth]{relative_risk_mean_across_ta.pdf}}
  \qquad
  \subfloat[\centering]{\includegraphics[width=.36\textwidth]{classifier_metrics_dist_across_ta.pdf}}
  \caption{
    \textbf{Statistical learning optimizes expected target-disease piar advancement rates using longitudinal evidence}.
    (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and Open Targets composite scores. 
    (b) Receiver operating characteristic (ROC) and average precision scores across the same 13 therapeutic areas with no limit on the number of rankings. 
    See Supplementary~Figure~\ref{fig:relative_risk_by_ta} for raw data underlying (a).
  }
  \label{fig:performance_across_ta}
\end{figure}

The relative risk estimates from the RDG model were also compared to univariate relative risk scores in Figure~\ref{fig:relative_risk_core_features}. This clearly implicates genetic features as being the most precise predictors, albeit with a relatively low frequency compared to others. Similar comparisons for target and disease specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}. These suggest that past clinical success for targets and diseases (indepent of pairings) are likely to be highly predictive of advancement as well, while genetic constraint and expression specificity of targets also exhibit modest but significant effects.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
  \caption{
    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs. The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years. All either features are assessed based on their existence. The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
  }
  \label{fig:relative_risk_core_features}
\end{figure}

In order to establish baseline levels of success and coverage across TD pairs, we refer to several of the univariate features in Figure~\ref{fig:relative_risk_core_features} throughout this study. Specifically, the \colorbox{Gainsboro}{omim} and \colorbox{Gainsboro}{ot\_genetics\_portal} features correspond to OMIM and GWAS baselines in the parlance of \cite{King2019-rc}, \cite{Nelson2015-eg} and \cite{Minikel2023.06.23.23291765}. We also use the \colorbox{Gainsboro}{eva} feature as a baseline for ClinVar support, which has discriminative capabilities somewhere between these two. The extent to which the RDG model meets or exceeds these benchmarks in both relative risk for advancement and TD pair coverage is illustrated in Figure~\ref{fig:relative_risk_by_limit}.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_by_limit.pdf}
  \caption{
    \textbf{RDG performance versus genetic support benchamrks across all therapeutic areas}. The RR estimates for each benchmark are based on the presence of the corresponding support across all TD pairs, and the number of pairs for which is present is represented by the horizontal bars extending horizontally from the y-axis. The bounds around the RDG RR estimates correspond to a 90\% confidence interval.
  }
  \label{fig:relative_risk_by_limit}
\end{figure}

One objective of this study is to determine if any model, e.g. RDG, could sort TD pairs with genetic support such that at least some portion of that sorted list had a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support alone. As a hypothetical example, we wanted to determine if the top 50 TD pairs out of 200 with GWAS support were more likely to advance than all 200 overall. We find that this goal is met and exceeded by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support alone at an expected rate of advancement exceeding that of the single source (respectively). This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult. See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities amongst undeveloped TD pairs.

\subsection{Sensitivity}

In order to validate the stability of our findings in Section~\ref{sec:results_performance}, we repeat this analysis across 18 different configurations listed in Supplementary~Table~\ref{tab:sensitivity_configurations}. This includes 3 separate versions of Open Targets, 3 choices for the year defining the split between training and evaluation data and 2 choices for the length of the minimum advancement window (in years).

We find that the mean RR values from the RDG model consistently exceed the OTS model in all configurations amongst the very highest rankings (N=10) and also exceed the OTS model in all configurations except for 1 for N between 20 and 60. This data is shown in Supplementary~Figure~\ref{fig:sensitivity_relative_risk}. The significance of these differences drops notably after N=40, which can be seen in the distribution of p-values from a Wilcoxon signed-rank test shown in Supplementary~Figure~\ref{fig:sensitivity_p_values}.

\subsection{Predictions}

Top predictions from the RDG model over the evaluation dataset are shown in Supplementary~Figure~\ref{fig:top_evaluation_predictions}. This figure demonstrates that there are a significant number of TD pairs with genetic support from many sources and that the combined effects of non-genetic features also substantially influence these rankings.

\subsection{Effects}

The coefficients learned by the RDG model, and the average effects they have across the evaluation dataset, are shown in Figure~\ref{fig:effect_sizes}. This model most highly prioritizes genetic signals that have the greatest coverage, i.e. associations from GWAS studies through the \colorbox{Gainsboro}{ot\_genetics\_portal} feature and associations from any curated clinical genetics source, i.e. EVA, Orphanet, UniProt, Genomics England, ClinGen and gene2phenotype, via the \colorbox{Gainsboro}{curated} \vspace*{0mm} feature.  Notably, literature and target/disease specific clinical features also have substantial effects, followed by indicators of animal evidence and target genetic constraint / expression specificity. Any features not shown were deflated to have no effect, which is possible in this model due to the non-negativity constraint. One such feature worth emphasizing is transcriptomic evidence from Expression Atlas. We found this somewhat surprising, but it is supported by arguments against transcript over/under expression as an indicator of genes that influence disease rather than the other way around \cite{PMID:34561431}.

\begin{figure}[!htb]
	\centering
  \captionsetup{width=.9\linewidth}
	\includegraphics[width=1\textwidth]{effect_sizes.pdf}
  \caption{
    \textbf{RDG model feature effects}. The \textbf{effect max} values are equivalent to RDG model coefficients for the corresponding feature while the \textbf{effect mean} values indicate average values of the product between the coefficient and a particular feature value, when that feature is present.
  }
	\label{fig:effect_sizes}
\end{figure}

It is worth noting that the discordance between the coefficients and the average feature effects of Figure~\ref{fig:effect_sizes} arises from both the frequency with which features exist as well as the distribution of their underlying scores. Scores for many clinical genetics features (e.g. OMIM, Genomics England, UniProt) are very frequently absent or close to 1. By comparison, scores for literature associations are typically far lower, even when limited only to cases where they exist, with a median value of 0.12 (mean=.23) in the evaluation data.

\subsection{Opportunities}
\label{sec:results_opportunities}

A common method for identifying druggable opportunities within a specific disease context involves first ranking TD pairs according to some prioritization methodology followed by filtering or reprioritizing those ranks based on knowledge of target tractability \cite{PMID:28356508,PMID:35401535,PMID:31253980}. We use a similar approach to identify tractable targets associated with TD pairs that have yet to enter clinical trials. To aid in interpreting this approach, we also draw on the results of Figure~\ref{fig:relative_risk_by_limit}. The data in this figure suggests thresholds for the RDG model that align to expected rates of advancement compared to several genetic support benchmarks. These thresholds are used to bucket undeveloped TD pairs before further bucketing them based on levels of tractability. The tractability buckets in Supplementary~Table~\ref{tab:tractability_buckets} provide \colorbox{Gainsboro}{HIGH}, \colorbox{Gainsboro}{MED}, and \colorbox{Gainsboro}{LOW} confidence ratings for each type of tractability evidence based on the priorities suggested in \cite{OTTractability}.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{opportunity_summary.png}
  \caption{
    \textbf{Present-day target-disease pair counts by tractability, likelihood of advancement and therapeutic area}. The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractactability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}. This corresponds to targets that have all been in clinical development already, except for the \textbf{OC} modality in which case it indicates that a target has been approved. 
  }
  \label{fig:opportunity_summary}
\end{figure}

Figure~\ref{fig:opportunity_summary} shows the distribution of TD pair counts for select buckets across therapeutic areas as well as across the current maximum phase reached for any one pair. We find that there are $\sim$2,400 small-molecule-enabled, $\sim$1,400 antibody-enabled, and 14 PROTAC-enabled TD pairs with a probability of advancement that is nearly 3x other TD pairs based on the EVA threshold RR=2.93 in Figure~\ref{fig:relative_risk_by_limit}. Top antibody-enabled pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.

\subsection{Algorithms}

The primary model for this analysis, RDG, is a constrained, L2-regularized linear regressor. Alternatives to this choice we explore include constrained and unconstrained gradient-boosted machine (GBM) models fit with the LightGBM \cite{LightGBM} algorithm as well as unconstrained L2-regularized linear regressors (i.e. RDG with no constraints). The performance of these alternatives, as quantified by relative risk estimates among top predictions, is illustrated in Figure~\ref{fig:relative_risk_model_features}.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
  \caption{
    \textbf{Performance by algorithm, constraint type and feature grouping}.  
  }
  \label{fig:relative_risk_model_features}
\end{figure}

We conclude from these results that 1) linear models outperform, or are at least not worse, than tree models when time is not used as a predictor and 2) non-negativity constraints in linear models result in greater performance. This is consistent with our expectation that all Open Targets scores should have monotonically increasing effects on the likliehood of success, since they were designed with this intent.

\subsection{Inflation}
\label{sec:results_inflation}

Like most studies of this kind, we assume a "closed-world" \cite{Paliwal2020-hr} over the space of target-disease pairs and any evidence between them. This means that we do not differentiate between evidence that an association for any one pair truly does \textbf{not} exist (or is too weak to be relevant under the omnigenic model \cite{PMID:28622505}), and the lack of any attempt to find that evidence in the first place. This also means that our estimate of the prognostic value for any one evidence source is subject to historical trends in biomedical research and the myriad ways that this research can be biased towards particular targets and diseases. We avoid attempting to comprehensively survey these biases in favor of offering an illustrative list of specific examples that are relevant in this study:

\begin{enumerate}
\item Mendelian randomization research is biased towards cardiovascular diseases as they have a disproportionate number of known, modifiable exposures \cite{PMID:36736292}
\item Putative protein interactions that do not result from genome-scale or otherwise unbiased assays result in an overrepresentation of successful drug targets in resources like STRING \cite{PMID:36370105}, thereby inflating the success of network expansion methods over these databases to identify such targets \cite{Sadler2023-xd}.
\item Transcript expression studies run in late-stage clinical trials for a single indication, e.g. \cite{PMID:27723281} linking SLE to IFN genes, are a degenerate indicator of advancement beyond earlier stage trials when the timing of this evidence is not accounted for.
\item Targets tested against more indications in clinical trials enrich for failures because the marginal cost of testing more indications decreases, but the evidence for these indications is often weaker \cite{PMID:33262371}.
\item Herding effects in pharma R\&D pipelines around particular drug targets are becoming increasingly clear over time \cite{PMID:37117303} and generate an excess of clinical evidence for those targets.
\end{enumerate}

We also note that the skew in basic drug target research towards those that already have rich annotations and well characterized molecular function \cite{PMID:29358745} as well as the disproportionate representation of particular target families in pharma R\&D pipelines \cite{PMID:27910877,PPR:PPR7029} and the fact that literature is well known to be biased away from negative results in general \cite{PMID:32893970} are all problematic.

While it is not possible to address all of these issues, we emphasize that there is a clear pattern across all of the examples in the list above in that they require \textbf{past} clinical successes and/or failures to arise in the first place. This suggests that accounting for when evidence first emerged would limit the extent of these problems. We do so in this study based solely on publication dates associated with any one piece of information linking target-disease pairs. This also offers a novel opportunity to attempt to quantify what kind of evidence suffers most from these biases. Figure~\ref{fig:evidence_inflation} presents results for this based on a relative risk statistic defined as:

\begin{equation}
  \frac{P(A | B)}{P(A | \neg B)}
\end{equation}

where:

\begin{itemize}
\item \(A\) is the event that evidence for a TD pair arises after its first early-stage (phase 1 or 2) trial rather than before
\item \(B\) is the event that a TD pair advances into late-stage trials (phase 3 or 4)
\end{itemize}

We refer to this as "inflation risk" so as not to confuse it with the relative risk statistic used in all other contexts and it can be more simply described as the fraction of TD pairs for which evidence arises \textbf{after} the beginning of an ultimately successful early-stage trial divided by that same fraction for TD pairs that do not advance to late-stage trials. The intuition for this statistic is that it will be higher if successful trials lead to the generation of evidence of a particular type, and it should be 1 in cases where the emergence of evidence is independent of clinical success. We also measure this potential lack of independence through the more commonly used Fisher's exact test, e.g. \cite{PMID:19725948}, and both are presented in Figure~\ref{fig:evidence_inflation}.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{evidence_inflation.pdf}
  \caption{
    \textbf{Clinical success drives evidence discovery}.  
  }
  \label{fig:evidence_inflation}
\end{figure}

We find that evidence from Reactome is the worst offender by this metric, implying that it often only arises for TD pairs after a certain level of clinical success has been attained. We also find that long-running aggregators/curators of published research often focused on individual diseases/phenotypes, like Expression Atlas, IMPC, CGS and Cancer Biomarkers exhibit this form of inflation as well.

Sources of genetic evidence appear to be much less inflated, or have too little data to reach significance. This is to be expected for GWAS evidence arising from genomewide-wide, phenome-wide biobank consortia, however much of historical GWAS evidence is not phenome-wide. More context on how much this is likely to matter comes from \cite{PMID:37612393} in which it was estimated that as few as 6\% of 500 FDA-approved targets for non-cancer drugs arose from programs highly motivated by pre-existing genetic support and that "the remaining 94\% were probably identified using conventional pharmacology, biochemistry or molecular biology approaches". We then speculate that if the initiation of new drug programs was not historically motivated highly by the existence of genetic support, then the incentives for pursing new genetic evidence based on clinical and commercial success are likely to be minimized. This, in conjunction with existing precedent \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:35804044} and our inflation results, ultimately led us to the use of genetic evidence without temporalization. In other words, we do not treat genetic evidence as longitudinal features like all others associated with TD pairs. A breakdown of which features are treated in which manner is provided in Supplementary~Table~\ref{tab:features}.

\section{Conclusion}

We have demonstrated that simple machine learning methods applied to longitudinal biomedical evidence from many sources can be used to predict clinical outcomes for combinations of drug targets and diseases, without knowledge of molecular properties or trial design details. We have also shown that these methods are more precise in the extremes of their predictions than composite, heuristic scores like those from Open Targets. They also outperform such baselines by more comprehensive, traditional measures of classifier performance; however, we find this less compelling and easier to accomplish than improving performance amongst the upper tail of the opportunities implied by the very highest predictions. This framework would also support the addition of new lines of evidence over time well as it is designed to automatically determine the relevance of any new information without intervention.  Lastly, we find that the space of present-day, undeveloped targets within a disease context that both exceed baseline levels of tractability and have a high predicted likelihood of clinical advancement is substantial. It is likely to grow quickly as well since the breadth of much of the underlying evidence is expanding rapidly \cite{PMID:33214558,PMID:36634672,PMID:31491408}.

\section{Discussion}
\label{sec:discussion}

\begin{itemize}
  \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
  \item Discuss the possibility to do prospective evaluation with OT snapshots
  \item Discuss how typical classifier metrics are inadequate for this problem
  \item Talk about how the inclusion of time greatly improves performance and suggests it may be possible to fill in this proxy with predictors for more specific early indicators of other reasons trials fail
  \item We are focusing on prioritizing amongst targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
  \item "There is evidence that a 9.6\% vs. 13.8\% success rate for drugs from phase 1 trials to approval may mean a \$480 million difference in the median research and development cost required to bring a new drug to the market (Wouters, McKee, and Luyten 2020)." \cite{PMID:34930919}
  \item From \cite{PMID:33262371}: "It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid."
  \item Add select tractability and DepMap features?
\end{itemize}

\section{Methods}
\label{sec:methods}

\begin{itemize}
  \item Discuss why all target prioritisation data fields are not used due to the potential leakage they may impose (e.g. target families, GO annotations, etc.)
  \item Describe how therapeutic areas were selected based on having at least 100 TD pairs with target-disease-specific evidence of any kind and with explicit omissions: ("biological\_process", "pregnancy or perinatal disease", "injury, poisoning or other complication", "pregnancy or perinatal disease", "medical procedure", "infectious disease", "animal disease")
  \item Training/evaluation results are limited by temporalization while the present-day predictions are not
  \item OMIM is defined as EVA associations with publications
  \item The "genetic\_association\_\_curated" field is a union of all genetic association sources other than "gene\_burden" and "ot\_genetics\_portal"
  \item Mention that the scores for a TD pairs in a year are the maximum score for that source, not the harmonic sum
  \item Mean imputation is used for target-specific features
  \item Mention specifics on RDG and GBM implementations, i.e. lightGBM and scikit-learn
  \item RDG is always trained on all features, but only applied to feature subsets where relevent
\end{itemize}

\section{Supplementary Material}

\input{assets/features_table.tex}

\input{assets/tractability_bucket_table.tex}

\input{assets/sensitivity_configurations.tex}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_by_ta.png}
  \caption{
    \textbf{Relative risk scores by method, benchmark and therapeutic area}.
    The \textbf{average} therapeutic area indicates mean values across all others except for \textbf{all}, which is an ungrouped estimate across all diseases regardless of therapeutic area.
  }
  \label{fig:relative_risk_by_ta}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{top_evaluation_predictions.png}
  \caption{
    \textbf{Top RDG model evaluation dataset predictions}.
    Feature contributions are shown as the product of their underlying values and the RDG coefficients. The \textbf{advanced} field indicates whether or not the associated TD pair advanced beyond phase 2 as of 2024.
  }
  \label{fig:top_evaluation_predictions}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{dataset_statistics.png}
  \caption{
    \textbf{Training and evaluation dataset summary statistics}.
  }
  \label{fig:dataset_statistics}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
  \caption{
    \textbf{Relative risk distributions across select therapeutic areas}.
    P-values are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR averages across therapeutic areas exceed OTS averages.
  }
  \label{fig:relative_risk_dist_across_ta}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}
  \caption{
    \textbf{Relative risk scores for target/disease features}.
    The features ending with \colorbox{Gainsboro}{qtl\_Q} denote binary indicators constructed from cases where the feature meets or exceeds quantile \colorbox{Gainsboro}{Q} of its distribution. The features ending with \colorbox{Gainsboro}{gte\_X} denote indicators for when the feature meets or exceeds a specific value \colorbox{Gainsboro}{X}.
  }
  \label{fig:relative_risk_static_features}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=.9\textwidth]{top_opportunity_predictions.png}
  \caption{
    \textbf{Top ranked undeveloped, tractable target-disease pairs}. The highest scoring TD pairs per the RDG model that have not entered clinical trials despite having a target that has been in trials of an antibody-based drug. All values shown other than \textbf{prediction} are raw feature values, unweighted by RDG coefficients.
  }
  \label{fig:top_opportunity_predictions}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{sensitivity_relative_risk.pdf}
  \caption{
    \textbf{Relative risk distributions across configurations in sensitivity analysis}.
    The distribution of the mean RR values displayed for a single configuration in Figure~\ref{fig:performance_across_ta} is shown here across 18 configurations.
  }
  \label{fig:sensitivity_relative_risk}
\end{figure}

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{sensitivity_p_values.pdf}
  \caption{
    \textbf{P-value distributions across configurations in sensitivity analysis}.
    The distribution of the p-values displayed for a single configuration in Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} is shown here across 18 configurations.
  }
  \label{fig:sensitivity_p_values}
\end{figure}

\bibliographystyle{unsrt}  
\bibliography{references}

\end{document}

### 2024-03-21 analysis.py

@@ -1,0 +1,3073 @@
+import functools
+import logging
+import math
+import re
+import os
+import sys
+import tempfile
+from collections.abc import Callable
+from dataclasses import dataclass
+from pathlib import Path
+from typing import Any, Concatenate, Literal, ParamSpec, TypeVar
+
+import fire
+import numpy as np
+import pandas as pd
+import psutil
+import xarray as xr
+from dotenv import load_dotenv
+from IPython.display import display
+from lightgbm import LGBMClassifier
+from pandas import DataFrame as PandasDataFrame
+from pandas import Series as PandasSeries
+from pandas.io.formats.style import Styler
+from pyspark.sql import DataFrame as SparkDataFrame
+from pyspark.sql import SparkSession, Window
+from pyspark.sql import functions as F
+from pyspark_gcs import get_gcs_enabled_config
+from scipy.stats import wilcoxon
+from scipy.stats.contingency import relative_risk
+from sklearn.base import BaseEstimator
+from sklearn.compose import ColumnTransformer
+from sklearn.discriminant_analysis import StandardScaler
+from sklearn.dummy import DummyClassifier
+from sklearn.impute import SimpleImputer
+from sklearn.linear_model import LogisticRegressionCV, Ridge
+from sklearn.pipeline import FunctionTransformer, Pipeline, make_pipeline
+from sklearn.preprocessing import MinMaxScaler
+from . import xrml
+
+logger = logging.getLogger(__name__)
+
+###############################################################################
+# Configuration
+###############################################################################
+
+SEED = 0
+FEATURES_PRIMARY_KEY = ["target_id", "disease_id", "year"]
+MIN_EVIDENCE_YEAR = 1990
+MAX_EVIDENCE_YEAR = 2024
+DEFAULT_MAX_TRAINING_TRANSITION_YEAR = 2015
+DEFAULT_CLINICAL_ADVANCEMENT_WINDOW = 2
+DATASETS = {
+    "23.12": (
+        default_tables := {
+            "evidence": "evidence",
+            "targets": "targets",
+            "diseases": "diseases",
+            "targetPrioritisation": "targetPrioritisation",
+            "associationByOverallDirect": "associationByOverallDirect",
+            "associationByOverallIndirect": "associationByOverallIndirect",
+        }
+    ),
+    "23.09": default_tables,
+    # Note: this typo in "Priorisation" is real for 23.06
+    "23.06": {**default_tables, **{"targetPrioritisation": "targetsPriorisation"}},
+}
+PRIMARY_TABLES = {
[…]

### 2024-03-21 tests/test_models.py

@@ -1,0 +1,76 @@
+import numpy as np
+import pandas as pd
+import pytest
+
+from pipelines.analyses.publications.clinical_progression_forecasting.public.analysis import (
+    get_imputer,
+    get_ridge_regression_pipeline,
+    make_pandas_pipeline,
+)
+
+
+def get_data(coefs: list[float], n: int = 100):
+    coef = np.array(coefs)
+    rng = np.random.RandomState(0)
+    n_samples = 10000
+    n_features = len(coef)
+    intercept = 1
+    X = pd.DataFrame(
+        rng.normal(loc=1, scale=3, size=(n_samples, n_features))
+    ).add_prefix("X")
+    y = pd.Series(intercept + (X.values * coef).sum(axis=1))
+    return X, y, pd.Series(coef, index=X.columns), intercept
+
+
+@pytest.mark.parametrize("positive", [True, False])
+def test_ridge_regression_coefficient_recovery(positive):
+    X, y, coefs, intercept = get_data([0, 1, 2] if positive else [-1, 0, 1], n=1_000)
+    feature_names = X.columns.tolist()
+    pipe = make_pandas_pipeline(
+        *get_ridge_regression_pipeline(
+            feature_names, positive=positive, add_imputer=False, add_scaler=False
+        )
+    ).fit(X, y)
+    np.testing.assert_allclose(pipe[-1].intercept_, intercept)
+    pd.testing.assert_series_equal(pipe[-1].coef_, coefs.astype(float))
+
+
+@pytest.mark.parametrize(
+    "inference_features", [None, ["X2"], ["X1", "X2"], ["X0", "X1", "X2"]]
+)
+def test_ridge_regression_inference_features(inference_features):
+    X, y, _, _ = get_data([-0.2, 0, 0.2])
+    feature_names = X.columns.tolist()
+    pipe = make_pandas_pipeline(
+        *get_ridge_regression_pipeline(
+            feature_names, inference_features=inference_features, add_imputer=False
+        )
+    ).fit(X, y)
+    expected = inference_features or feature_names
+    actual = pipe[-1].explain(pipe[:-1].transform(X)).columns.tolist()
+    assert actual == expected
+
+
+def test_imputer():
+    X = pd.DataFrame(
+        {
+            "target__static_1": [1, 1, 2, 2, 3, 3],
+            "target__static_2": [1, 1, 2, 2, None, None],
[…]

### 2024-03-21 xrml.py

@@ -1,0 +1,1329 @@
+import inspect
+import json
+import logging
+import os
+import warnings
+from collections.abc import Callable, Hashable, Sequence
+from dataclasses import dataclass, replace
+from dataclasses import field as dataclass_field
+from typing import Any, Literal
+
+import dill
+import fsspec
+import numpy as np
+import numpy.typing as npt
+import pandas as pd
+import shap
+import tqdm
+import xarray as xr
+from fsspec import AbstractFileSystem, get_filesystem_class
+from fsspec.utils import get_protocol
+from scipy.stats import kendalltau
+from shap import Explanation
+from shap.maskers import Independent
+from sklearn.base import BaseEstimator, clone
+from sklearn.exceptions import ConvergenceWarning
+from sklearn.metrics import (
+    average_precision_score,
+    brier_score_loss,
+    explained_variance_score,
+    f1_score,
+    log_loss,
+    matthews_corrcoef,
+    mean_absolute_error,
+    mean_squared_error,
+    ndcg_score,
+    precision_score,
+    r2_score,
+    recall_score,
+    roc_auc_score,
+)
+from sklearn.model_selection import GroupKFold, StratifiedGroupKFold, StratifiedKFold
+from sklearn.pipeline import Pipeline
+from sklearn.utils.multiclass import type_of_target
+from xarray import DataArray, Dataset
+from pandas import DataFrame as PandasDataFrame
+
+logger = logging.getLogger(__name__)
+
+
+@dataclass
+class Model:
+    slug: str
+    """ Human-readable unique model identifier, e.g. 'logreg' """
+
+    name: str
+    """ Display name, e.g. 'Logistic Regression' """
+
+    estimator: BaseEstimator
+    """ Estimator instance """
+
+    properties: dict[str, Any] = dataclass_field(default_factory=lambda: {})
+    """ Arbitrary properties/metadata for the model """
+
+
+#######################
+# Xarray CV Functions #
+#######################
+
+
+def split(ds: Dataset) -> tuple[DataArray | None, DataArray | None]:
+    """Break dataset into features and outcomes for modeling"""
+    X = None
+    if "feature" in ds:
+        if ds.dims["features"] < 1:
+            raise ValueError("Must provide at least one feature")
+        X = ds["feature"].transpose("index", "features")
+        assert X.ndim == 2
+
+    Y = None
+    if "outcome" in ds:
+        if ds.dims["outcomes"] < 1:
[…]

### 2024-03-27 paper/paper.tex
Fix spelling and grammar errors

@@
 This study examines the extent to which the outcomes of clinical trials can be predicted based on longitudinal properties of drug targets and diseases alone.
-We find that this is possible by comparing the historical performance of model-based target-disease pair priortiziation methods to common baselines.
+We find that this is possible by comparing the historical performance of model-based target-disease pair prioritization methods to common baselines.
 Our primary objective is to demonstrate that statistical learning can effectively optimize such methods with no loss of interpretability.
@@
 It is also possible to expand the breadth of this genetic support to more targets/diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}.
-To our knolwedge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates.
+To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates.
 This is not a focus of this work as we aim, instead, to establish a framework for identifying targets/diseases with the very highest possible likelihood of success first.
 This is accomplished by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in a simple statistical modeling framework.
-This can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret.
+It can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret.
 We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence from many sources, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.
@@
 
-Addressing such problems is common, but not ubuquitous, in studies that attempt to predict clincal trial outcomes using temporalized predictors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}.
+Addressing such problems is common, but not ubiquitous, in studies that attempt to predict clinical trial outcomes using temporalized predictors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}.
 The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets/diseases, trial sponsor track records, eventual patient enrollment, etc.
 constitute clear information leaks otherwise.
-This is discussed more in \cite{PMID:37483175} which notes several studies that do not account for this problem, and presents both "quasi-prospective" as well as true "prospective" results.
-The difference between these two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until years later.
+This is discussed more in \cite{PMID:37483175} which notes several studies that do not account for this problem, and presents "quasi-prospective" as well as true "prospective" results.
+The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until years later.
 Nomenclature for these formulations is conflicting though, where this definition of a "quasi-prospective" design is deemed entirely prospective in some cases, e.g.
@@
 The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete.
-While the average phase 2 trial duration may be as lower as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}.
+While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}.
 This means a substantial fraction of outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation.
@@
 A subset of therapeutic areas was used for this analysis and this is reflected in both the mean estimates and distributions of Figure~\ref{fig:performance_across_ta}.
-A full list of these can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta}, which also shows the the RR estimates used in any summary statistics across therapeutic areas.
+A full list of these can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta}, which also shows the RR estimates used in any summary statistics across therapeutic areas.
 These were selected based on criteria described in Section~\ref{sec:methods}.
@@
 Similar comparisons for target and disease specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}.
-These suggest that past clinical success for targets and diseases (indepent of pairings) are likely to be highly predictive of advancement as well, while genetic constraint and expression specificity of targets also exhibit modest but significant effects.
+These suggest that past clinical success for targets and diseases (independent of pairings) are likely to be highly predictive of advancement as well, while genetic constraint and expression specificity of targets also exhibit modest but significant effects.
 
@@
 The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
-All either features are assessed based on their existence.
+All other features are assessed based on their existence.
 The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
@@
   \caption{
-    \textbf{RDG performance versus genetic support benchamrks across all therapeutic areas}.
+    \textbf{RDG performance versus genetic support benchmarks across all therapeutic areas}.
 The RR estimates for each benchmark are based on the presence of the corresponding support across all TD pairs, and the number of pairs for which is present is represented by the horizontal bars extending horizontally from the y-axis.
@@
 This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult.
-See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities amongst undeveloped TD pairs.
+See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs.
 
@@
 
-We find that the mean RR values from the RDG model consistently exceed the OTS model in all configurations amongst the very highest rankings (N=10) and also exceed the OTS model in all configurations except for 1 for N between 20 and 60.
+We find that the mean RR values from the RDG model consistently exceed the OTS model in all configurations among the very highest rankings (N=10) and also exceed the OTS model in all configurations except for 1 for N between 20 and 60.
 This data is shown in Supplementary~Figure~\ref{fig:sensitivity_relative_risk}.
@@
 
-It is worth noting that the discordance between the coefficients and the average feature effects of Figure~\ref{fig:effect_sizes} arises from both the frequency with which features exist as well as the distribution of their underlying scores.
+It is worth noting that the discordance between the coefficients and the average feature effects of Figure~\ref{fig:effect_sizes} arises from both the frequency with which features exist and the distribution of their underlying scores.
 Scores for many clinical genetics features (e.g.
@@
     \textbf{Present-day target-disease pair counts by tractability, likelihood of advancement and therapeutic area}.
-The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractactability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}.
+The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}.
 This corresponds to targets that have all been in clinical development already, except for the \textbf{OC} modality in which case it indicates that a target has been approved.
@@
 We conclude from these results that 1) linear models outperform, or are at least not worse, than tree models when time is not used as a predictor and 2) non-negativity constraints in linear models result in greater performance.
-This is consistent with our expectation that all Open Targets scores should have monotonically increasing effects on the likliehood of success, since they were designed with this intent.
+This is consistent with our expectation that all Open Targets scores should have monotonically increasing effects on the likelihood of success, since they were designed with this intent.
 
@@
 
-While it is not possible to address all of these issues, we emphasize that there is a clear pattern across all of the examples in the list above in that they require \textbf{past} clinical successes and/or failures to arise in the first place.
+While it is not possible to address all of these issues, we emphasize that there is a clear pattern across the examples in the list above in that they require \textbf{past} clinical successes and/or failures to arise in the first place.
 This suggests that accounting for when evidence first emerged would limit the extent of these problems.
@@
 
-We refer to this as "inflation risk" so as not to confuse it with the relative risk statistic used in all other contexts and it can be more simply described as the fraction of TD pairs for which evidence arises \textbf{after} the beginning of an ultimately successful early-stage trial divided by that same fraction for TD pairs that do not advance to late-stage trials.
+We refer to this as "inflation risk" so as not to confuse it with the relative risk statistic used in all other contexts, and it can be more simply described as the fraction of TD pairs for which evidence arises \textbf{after} the beginning of an ultimately successful early-stage trial divided by that same fraction for TD pairs that do not advance to late-stage trials.
 The intuition for this statistic is that it will be higher if successful trials lead to the generation of evidence of a particular type, and it should be 1 in cases where the emergence of evidence is independent of clinical success.
@@
 Sources of genetic evidence appear to be much less inflated, or have too little data to reach significance.
-This is to be expected for GWAS evidence arising from genomewide-wide, phenome-wide biobank consortia, however much of historical GWAS evidence is not phenome-wide.
+This is to be expected for GWAS evidence arising from genome-wide, phenome-wide biobank consortia, however much of historical GWAS evidence is not phenome-wide.
 More context on how much this is likely to matter comes from \cite{PMID:37612393} in which it was estimated that as few as 6\% of 500 FDA-approved targets for non-cancer drugs arose from programs highly motivated by pre-existing genetic support and that "the remaining 94\% were probably identified using conventional pharmacology, biochemistry or molecular biology approaches".
@@
 We have also shown that these methods are more precise in the extremes of their predictions than composite, heuristic scores like those from Open Targets.
-They also outperform such baselines by more comprehensive, traditional measures of classifier performance; however, we find this less compelling and easier to accomplish than improving performance amongst the upper tail of the opportunities implied by the very highest predictions.
+They also outperform such baselines by more comprehensive, traditional measures of classifier performance; however, we find this less compelling and easier to accomplish than improving performance among the upper tail of the opportunities implied by the very highest predictions.
 This framework would also support the addition of new lines of evidence over time well as it is designed to automatically determine the relevance of any new information without intervention.
@@
   \item Talk about how the inclusion of time greatly improves performance and suggests it may be possible to fill in this proxy with predictors for more specific early indicators of other reasons trials fail
-  \item We are focusing on prioritizing amongst targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
+  \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
   \item "There is evidence that a 9.6\% vs.
@@
   \item Mention specifics on RDG and GBM implementations, i.e. lightGBM and scikit-learn
-  \item RDG is always trained on all features, but only applied to feature subsets where relevent
+  \item RDG is always trained on all features, but only applied to feature subsets where relevant
 \end{itemize}
@@
     Feature contributions are shown as the product of their underlying values and the RDG coefficients.
-The \textbf{advanced} field indicates whether or not the associated TD pair advanced beyond phase 2 as of 2024.
+The \textbf{advanced} field indicates whether the associated TD pair advanced beyond phase 2 as of 2024.
   }

### 2024-03-28 analysis.py
Refactor algorithm and feature set figure

@@ -35,7 +35,7 @@ from sklearn.impute import SimpleImputer
 from sklearn.linear_model import LogisticRegressionCV, Ridge
 from sklearn.pipeline import FunctionTransformer, Pipeline, make_pipeline
 from sklearn.preprocessing import MinMaxScaler
-from . import xrml
+import xrml
 
 logger = logging.getLogger(__name__)
 
@@ -2507,8 +2507,8 @@ def prepare_classification_metrics_by_therapeutic_area(
                 df.assign(clipped=False)
             ) if ap_max_value is None else (
                 df
-                .assign(clipped=lambda df: np.where(df["metric"] == "average precision", df["value"] > .3, False))
-                .assign(value=lambda df: np.where(df["metric"] == "average precision", df["value"].clip(0, .3), df["value"]))
+                .assign(clipped=lambda df: np.where(df["metric"] == "average precision", df["value"] > ap_max_value, False))
+                .assign(value=lambda df: np.where(df["metric"] == "average precision", df["value"].clip(0, ap_max_value), df["value"]))
             )
         ))
     )
@@ -2565,9 +2565,10 @@ def prepare_relative_risk_by_therapeutic_area(
     # fmt: off
     return (
         relative_risk
-        .pipe(lambda df: df[df["is_primary_benchmark_feature"] | df["is_primary_model_feature"]])
-        # Fill limit before using in grouping and invert fill after
+        .pipe(lambda df: df[df["is_primary_benchmark_feature"] | df["is_model_feature"]])
+        # Fill limit before using in grouping and invert fill after (it is dropped otherwise as Int64)
         .assign(limit=lambda df: df["limit"].fillna(0))
+        # Compute summaries across therapeutic areas
         .pipe(lambda df: pd.concat([
             df,
             df
@@ -2576,6 +2577,7 @@ def prepare_relative_risk_by_therapeutic_area(
             .agg(**{
                 **{
                     "relative_risk": ("relative_risk", "mean"),
+                    "n_therapeutic_areas": ("therapeutic_area_name", "nunique")
                 },
                 **{
                     c: (c, "first")
@@ -2597,7 +2599,21 @@ def prepare_relative_risk_by_therapeutic_area(
             how='left'
         ))
         .assign(limit=lambda df: df["limit"].where(df["limit"] > 0).astype('Int64'))
-        .pipe(assert_condition, lambda df: (~df["is_primary_model_feature"] | df["p_value"].notnull()).all())
[…]

### 2024-03-29 paper/paper.tex
Fix spelling and grammar errors
Finish refactoring of Figure 2

@@
   \caption{
-    \textbf{Statistical learning optimizes expected target-disease piar advancement rates using longitudinal evidence}.
+    \textbf{Statistical learning optimizes expected target-disease pair advancement rates using longitudinal evidence}.
     (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and Open Targets composite scores. 
@@
   \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
+  \includegraphics[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
   \caption{
-    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
-    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs.
-The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
-All other features are assessed based on their existence.
-The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
+    \textbf{Performance of ...}.
   }
-  \label{fig:relative_risk_core_features}
+  \label{fig:performance_metric_mean_across_ta}
 \end{figure}
@@
   \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
+  \caption{
+    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
+    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs.
+The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
+All other features are assessed based on their existence.
+The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
+  }
+  \label{fig:relative_risk_core_features}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \captionsetup{width=.9\linewidth}
   \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}

### 2024-03-29 analysis.py

@@ -2707,6 +2707,109 @@ def combine_average_therapeutic_metrics(
         .assign(model_algorithm=lambda df: df["model_name"].str.split("__").str[0].fillna(""))
         .assign(model_features=lambda df: df["model_name"].str.split("__").str[1].fillna(""))
         .assign(model_constraint=lambda df: df["model_name"].str.split("__").str[2].fillna(""))
+        .pipe(assert_condition, lambda df: not df[["model_name", "metric"]].duplicated().any())
+    )
+    # fmt: on
+
+
+def prepare_average_therapeutic_area_metrics(
+    average_therapeutic_metrics: PandasDataFrame,
+) -> PandasDataFrame:
+    metrics = [
+        "rr@010",
+        "rr@020",
+        "rr@030",
+        "rr@040",
+        "rr@050",
+        "rr@100",
+        "average_precision",
+        "roc_auc",
+    ]
+
+    feature_group = "feature ablations\n([+] models only)"
+    algorithms_group = "model algorithms\n(evidence only)"
+    df = average_therapeutic_metrics.pipe(
+        lambda df: df[df["has_all_primary_therapeutic_areas"]]
+    ).pipe(lambda df: df[df["metric"].isin(metrics)])
+    # fmt: off
+    return (
+        pd.concat(
+            [
+                (
+                    df
+                    .pipe(lambda df: df[df["model_algorithm"].isin(["rdg", "ots"])])
+                    .pipe(lambda df: df[df["model_features"].isin(["all", "no_time", "no_tgc"])])
+                    .pipe(lambda df: df[(df["model_algorithm"] == "ots") | (df["model_constraint"] == "positive")])
+                    .assign(
+                        model=lambda df: (
+                            df["model_algorithm"].str.upper() + 
+                            np.where(
+                                df["model_algorithm"] == "ots", "",
+                                df["model_features"].map({
+                                    "all": "-T",
+                                    "no_time": "",
+                                    "no_tgc": "-X",
+                                })
+                            )
+                        )
+                    )
+                    .assign(group=feature_group)
+                ),
+                (
+                    df
+                    .pipe(lambda df: df[df["model_algorithm"].isin(["gbm", "rdg", "ots"])])
+                    .pipe(lambda df: df[
+                        (df["model_features"] == "no_time") | 
+                        (df["model_algorithm"] == "ots")
+                    ])
+                    .assign(
+                        model=lambda df: (
+                            df["model_algorithm"].str.upper()
[…]

### 2024-03-30 paper/paper.tex
Minor figure and caption improvements

@@
 
-Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study -- a constrained, linear L2-regularized model referred to from here on as "RDG" -- ranks TD pairs by comparison to a composite score from Open Targets ("OTS").
+Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study -- a constrained, linear L2-regularized model referred to from here on as "RDG" (short for ridge regression) -- ranks TD pairs by comparison to a composite score from Open Targets ("OTS").
 The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:
@@
 As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available.
-They are, however, a useful performance ceiling towards which future work might build towards.
+They are, however, a useful performance ceiling towards which future work could build.
 
@@
 These were selected based on criteria described in Section~\ref{sec:methods}.
-Lastly, Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} shows RR distributions by model with significances of their differences.
+Lastly, Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} shows RR distributions by model with the significance of their differences.
 
@@
   \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
+  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
   \caption{
-    \textbf{Performance of ...}.
+    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
+    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs.
+The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
+All other features are assessed based on their existence.
+The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
   }
-  \label{fig:performance_metric_mean_across_ta}
+  \label{fig:relative_risk_core_features}
 \end{figure}
@@
 See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs.
+
+\subsection{Models}
+
+The primary model for this analysis, RDG, is a constrained, L2-regularized linear regressor.
+Alternatives to this choice we explore include constrained and unconstrained gradient-boosted machine (GBM) models fit with the LightGBM \cite{LightGBM} algorithm as well as unconstrained L2-regularized linear regressors (i.e.
+RDG with no constraints).
+The performance of these alternatives, as quantified by relative risk estimates among top predictions, is illustrated in Figure~\ref{fig:performance_metric_mean_across_ta}.
+
+\begin{figure}[!htb]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
+  \caption{
+    \textbf{Performance of ...}.
+  }
+  \label{fig:performance_metric_mean_across_ta}
+\end{figure}
+
+We conclude from these results that 1) linear models outperform, or are at least not worse, than tree models when time is not used as a predictor and 2) non-negativity constraints in linear models result in greater performance.
+This is consistent with our expectation that all Open Targets scores should have monotonically increasing effects on the likelihood of success, since they were designed with this intent.
 
@@
   \caption{
-    \textbf{Present-day target-disease pair counts by tractability, likelihood of advancement and therapeutic area}.
+    \textbf{Present-day target-disease pair counts by stage, likelihood of advancement and tractability}.
 The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}.
@@
 Top antibody-enabled pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.
-
-\subsection{Algorithms}
-
-The primary model for this analysis, RDG, is a constrained, L2-regularized linear regressor.
-Alternatives to this choice we explore include constrained and unconstrained gradient-boosted machine (GBM) models fit with the LightGBM \cite{LightGBM} algorithm as well as unconstrained L2-regularized linear regressors (i.e.
-RDG with no constraints).
-The performance of these alternatives, as quantified by relative risk estimates among top predictions, is illustrated in Figure~\ref{fig:relative_risk_model_features}.
-
-\begin{figure}[!htb]
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
-  \caption{
-    \textbf{Performance by algorithm, constraint type and feature grouping}.  
-  }
-  \label{fig:relative_risk_model_features}
-\end{figure}
-
-We conclude from these results that 1) linear models outperform, or are at least not worse, than tree models when time is not used as a predictor and 2) non-negativity constraints in linear models result in greater performance.
-This is consistent with our expectation that all Open Targets scores should have monotonically increasing effects on the likelihood of success, since they were designed with this intent.
 
@@
   \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
-  \caption{
-    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
-    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs.
-The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
-All other features are assessed based on their existence.
-The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
-  }
-  \label{fig:relative_risk_core_features}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \captionsetup{width=.9\linewidth}
   \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}
@@
   \label{fig:relative_risk_static_features}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
+  \caption{
+    \textbf{Performance by algorithm, constraint type and feature grouping}.  
+  }
+  \label{fig:relative_risk_model_features}
 \end{figure}

### 2024-03-30 analysis.py

@@ -14,7 +14,7 @@ import fire
 import numpy as np
 import pandas as pd
 import psutil
-import xarray as xr
+from xarray import Dataset
 from dotenv import load_dotenv
 from IPython.display import display
 from lightgbm import LGBMClassifier
@@ -989,7 +989,7 @@ def get_feature_statistics(
 
 
 def get_feature_relative_risk(
-    ds: xr.Dataset,
+    ds: Dataset,
     model_names: list[str] | None = None,
     model_limits: list[int] | None = None,
     confidence: float = 0.9,
@@ -1201,10 +1201,10 @@ def enrich_feature_relative_risk(
 
 
 def get_classification_metrics_by_therapeutic_area(
-    ds: xr.Dataset, therapeutic_areas: PandasDataFrame
+    ds: Dataset, therapeutic_areas: PandasDataFrame
 ) -> PandasDataFrame:
     def _metrics(
-        ds: xr.Dataset, therapeutic_area_id: str, therapeutic_area_name: str
+        ds: Dataset, therapeutic_area_id: str, therapeutic_area_name: str
     ) -> PandasDataFrame:
         disease_ids = (
             therapeutic_areas.pipe(
@@ -1243,7 +1243,7 @@ def get_classification_metrics_by_therapeutic_area(
 
 
 def get_relative_risk_thresholds(
-    relative_risk: PandasDataFrame, ds: xr.Dataset
+    relative_risk: PandasDataFrame, ds: Dataset
 ) -> PandasDataFrame:
     models = (
         relative_risk.pipe(lambda df: df[df["is_primary_model_feature"]])["model_name"]
@@ -1340,7 +1340,7 @@ def get_relative_risk_thresholds(
 
 
 def get_relative_risk_by_therapuetic_area(
-    ds: xr.Dataset,
+    ds: Dataset,
     therapeutic_areas: PandasDataFrame,
     model_names: list[str],
     model_limits: list[int] | None = None,
@@ -1536,7 +1536,7 @@ def get_modeling_features(
     return result
 
 
-def create_modeling_dataset(features: pd.DataFrame) -> xr.Dataset:
+def create_modeling_dataset(features: pd.DataFrame) -> Dataset:
     index = ["target_id", "disease_id"]
     descriptors = [c for c in features if "__" not in c and c not in index]
     assert len([c for c in features if "__outcome__" in c]) == 1
[…]

### 2024-03-31 paper/paper.tex
Configure subfigure label positions

@@
 \usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
 \usepackage{nicefrac}       % compact symbols for 1/2, etc.
 \usepackage{microtype}      % microtypography
-\usepackage{lipsum}
 \usepackage{fancyhdr}       % header
@@
 \usepackage{subfig}
+\usepackage{caption}
 \usepackage[section]{placeins}
 \usepackage{setspace}
+\usepackage{overpic}
 
@@
   \centering
-  \captionsetup{width=.9\linewidth}
-  \subfloat[\centering]{\includegraphics[width=.58\textwidth]{relative_risk_mean_across_ta.pdf}}
+  \captionsetup[subfigure]{labelformat=empty,width=.9\linewidth}
+  \subfloat[\centering]{
+    \begin{overpic}[width=.57\textwidth]{relative_risk_mean_across_ta.pdf}
+      \put(50, -3) {(a)}
+    \end{overpic}
+  }
   \qquad
-  \subfloat[\centering]{\includegraphics[width=.36\textwidth]{classifier_metrics_dist_across_ta.pdf}}
+  \subfloat[\centering]{
+    \begin{overpic}[width=.36\textwidth]{classifier_metrics_dist_across_ta.pdf}
+      \put(27, -3) {(b)}
+    \end{overpic}
+  }
   \caption{
@@
   \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
+  \begin{overpic}[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
+    \put(84.7, 38.2) {(a)}
+    \put(84.7, 22.5) {(b)}
+  \end{overpic}
   \caption{

### 2024-04-01 paper/paper.tex
Finish exposition for feature ablations and algorithms

@@
 \usepackage{subfig}
+\usepackage{float}
 \usepackage{caption}
@@
 All results to follow consider only advancement beyond phase 2 due to limitations described in Section~\ref{sec:discussion}.
-This binary outcome is then predicted based on a list of features shown in Supplementary~Table~\ref{tab:features}.
+This binary outcome is then predicted based on a list of features shown in Table~\ref{tab:features}.
 Information for each of these features is only used when it was published before the year \textbf{prior} to the first phase 2 trial observed, with an exception for genetic evidence discussed in Section~\ref{sec:results_inflation}.
@@
 
-The predictors used consist of 27 target-disease pair features, 5 target-specific features and 1 disease-specific feature.
-These are listed in Supplementary~Table~\ref{tab:features}.
+The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor.
+These are listed in Table~\ref{tab:features}.
 The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods that do not favor well-studied or drugged targets/diseases.
@@
 \subsection{Models}
+\label{sec:models}
 
 We train a variety of models including constrained and unconstrained linear and tree models.
-The constrained variants of these models force effects of all features to increase monotonically.
-This is possible with no feature transformations because all scores in Open Targets are constructed such that higher scores are presumed to be advantageous.
+The constrained variants of these models force effects of all features to increase monotonically, i.e.
+all effects are constrained to be non-negative.
+This is possible with no underlying feature transformations because all scores in Open Targets are constructed such that higher scores are presumed to be advantageous.
+
 We also apply these models to our evaluation dataset using several feature ablations in order to assess the value of groups of related features.
+We refer to a "core" feature set consisting of all features listed in Table~\ref{tab:features} except for the sole feature capturing the time since a target-disease pair first entered phase 2 trials (i.e.
+\colorbox{Gainsboro}{target\_disease\_\_time\_\_transition}).
+Combinations of learning algorithms for the models and the feature sets to which they are applied are referred to using the following convention:
 
-In order to compare these models to an Open Targets composite score, we use an equally weighted sum of all scores except for those assigned lower weights in \cite{OTweights}.
-Scores from these sources are multiplied by the corresponding weight before being summed.
+\begin{itemize}
+  \item \textbf{RDG}: Constrained, L2-regularized linear regressor (a.k.a. "Ridge regressor") fit with all core features
+  \item \textbf{RDG-T}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
+  \item \textbf{RDG-X}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
+  \item \textbf{GBM}: Constrained, gradient-boosted machine with fit with all core features
+  \item \textbf{GBM-T}: \textbf{GBM} fit with all features instead of only core features
+  \item \textbf{OTS}: Open Targets composite score
+\end{itemize}
+
+The omitted human clinical and genetic evidence features for the RDG-X model are all of those in Table~\ref{tab:features} with the midfix "clinical" or "genetic\_association".
+We omit these features specifically because they are known or expected to comprise good predictors of human clinical success, so their exclusion examines the extent to which only literature and animal model evidence along with target-specific properties accomplish this task.
+
+In order to compare these models to an Open Targets composite score (OTS), we use an equally weighted sum of all scores from individual sources except for those assigned lower weights in \cite{OTweights}.
+Scores from these sources are multiplied by the corresponding weight before being summed and only the TD-specific features of Table~\ref{tab:features} are used.
+Neither the target/disease specific features nor the time since phase 2 transition feature are included in this calculation.
 
@@
 
-Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study -- a constrained, linear L2-regularized model referred to from here on as "RDG" (short for ridge regression) -- ranks TD pairs by comparison to a composite score from Open Targets ("OTS").
+\subsubsection{Open Targets comparison}
+
+Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study, RDG, ranks TD pairs by comparison to a composite score from Open Targets, OTS.
 The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:
@@
 
-The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it also uses time since the phase 2 transition as a predictive factor as well as all others.
-We observe that the use of this information almost universally improves standard performance metrics like ROC (receiver operating characteristic) and AP (average precision), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed.
+The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it uses time since the phase 2 transition as a predictive factor in addition to all others.
+We observe that the use of this information almost universally improves standard performance metrics like receiver operating characteristic (ROC) and average precision (AP), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed.
 In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support.
-This is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion}).
+More implications of this and opportunities it may imply are discussed in Section~\ref{sec:discussion}.
 As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available.
-They are, however, a useful performance ceiling towards which future work could build.
-
-A subset of therapeutic areas was used for this analysis and this is reflected in both the mean estimates and distributions of Figure~\ref{fig:performance_across_ta}.
-A full list of these can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta}, which also shows the RR estimates used in any summary statistics across therapeutic areas.
-These were selected based on criteria described in Section~\ref{sec:methods}.
-Lastly, Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} shows RR distributions by model with the significance of their differences.
+They do, however, present a useful performance ceiling towards which future work might build.
 
@@
   \centering
-  \captionsetup[subfigure]{labelformat=empty,width=.9\linewidth}
+  \captionsetup{width=.9\linewidth}
+  \captionsetup[subfigure]{labelformat=empty}
   \subfloat[\centering]{
@@
   \caption{
-    \textbf{Statistical learning optimizes expected target-disease pair advancement rates using longitudinal evidence}.
-    (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and Open Targets composite scores. 
+    \textbf{Performance compared to Open Targets composite scores}.
+    (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and OTS (Open Targets composite scores). 
     (b) Receiver operating characteristic (ROC) and average precision scores across the same 13 therapeutic areas with no limit on the number of rankings. 
@@
 
-The relative risk estimates from the RDG model were also compared to univariate relative risk scores in Figure~\ref{fig:relative_risk_core_features}.
-This clearly implicates genetic features as being the most precise predictors, albeit with a relatively low frequency compared to others.
-Similar comparisons for target and disease specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}.
-These suggest that past clinical success for targets and diseases (independent of pairings) are likely to be highly predictive of advancement as well, while genetic constraint and expression specificity of targets also exhibit modest but significant effects.
+We also note that Figure~\ref{fig:performance_across_ta} presents average RR estimates drawn across a subset of therapeutic areas, and the criteria used to select them is described more in Section~\ref{sec:methods}.
+A full list of therapeutic areas meeting these criteria can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta} along with the RR estimates summarized in Figure~\ref{fig:performance_across_ta}.
+Furthermore, a comparison of the distribution of these estimates by model is presented in Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} along with the statistical significance of their differences.
 
-\begin{figure}[!htb]
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
-  \caption{
-    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
-    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs.
-The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
-All other features are assessed based on their existence.
-The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
-  }
-  \label{fig:relative_risk_core_features}
-\end{figure}
+\subsubsection{Genetic benchmark comparison}
 
-In order to establish baseline levels of success and coverage across TD pairs, we refer to several of the univariate features in Figure~\ref{fig:relative_risk_core_features} throughout this study.
-Specifically, the \colorbox{Gainsboro}{omim} and \colorbox{Gainsboro}{ot\_genetics\_portal} features correspond to OMIM and GWAS baselines in the parlance of \cite{King2019-rc}, \cite{Nelson2015-eg} and \cite{Minikel2023.06.23.23291765}.
-We also use the \colorbox{Gainsboro}{eva} feature as a baseline for ClinVar support, which has discriminative capabilities somewhere between these two.
-The extent to which the RDG model meets or exceeds these benchmarks in both relative risk for advancement and TD pair coverage is illustrated in Figure~\ref{fig:relative_risk_by_limit}.
+In order to establish baseline levels of success and coverage across TD pairs, we examine ranking performance in comparison to well established, univariate indicators of genetic support in Figure~\ref{fig:relative_risk_by_limit}.
+This figure presents OMIM and GWAS baselines, in the parlance of \cite{King2019-rc}, \cite{Nelson2015-eg} and \cite{Minikel2023.06.23.23291765}, as well as an intermediate baseline from the European Variation Archive (EVA) \cite{PMID:34718739} containing evidence predominantly from ClinVar \cite{PMID:24234437}.
+
+One key objective of this study is to determine if any model, e.g.
+RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeding what is expected from any one source of genetic support alone.
+We find that this goal is met and exceeded by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support alone at an expected rate of advancement exceeding that of the single source (respectively).
+This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult.
+See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model.
+This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.
 
@@
   \caption{
-    \textbf{RDG performance versus genetic support benchmarks across all therapeutic areas}.
+    \textbf{Performance compared to genetic support benchmarks}.
 The RR estimates for each benchmark are based on the presence of the corresponding support across all TD pairs, and the number of pairs for which is present is represented by the horizontal bars extending horizontally from the y-axis.
-The bounds around the RDG RR estimates correspond to a 90\% confidence interval.
+The bounds around the RDG RR estimates correspond to a Katz 90\% confidence interval.
   }
@@
 
-One objective of this study is to determine if any model, e.g.
-RDG, could sort TD pairs with genetic support such that at least some portion of that sorted list had a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support alone.
-As a hypothetical example, we wanted to determine if the top 50 TD pairs out of 200 with GWAS support were more likely to advance than all 200 overall.
-We find that this goal is met and exceeded by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support alone at an expected rate of advancement exceeding that of the single source (respectively).
-This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult.
-See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs.
+We also note that Supplementary~Figure~\ref{fig:relative_risk_core_features} shows confidence intervals for each of the genetic benchmarks of Figure~\ref{fig:relative_risk_by_limit} in isolation, as well as all other target-disease-specific evidence sources, in addition to confidence intervals for the RDG model at various top ranking cutoffs.
+Similar comparisons for target-specific and disease-specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}.
+These findings suggest that 1) genetic support for TD pairs is highly predictive but rare, 2) human clinical support for targets and diseases in isolation is also predictive while being more common and 3) constraint and expression specificity of targets exhibit modest but significant effects.
+While only human clinical evidence appears to have a prognostic value rivaling that of genetic support when considered on a univariate basis, the combined influence of non-genetic, non-clinical evidence is examined in Section~\ref{sec:model_comparison} where a model using only this information, RDG-X, still outperforms Open Targets composite scores by all measures.
 
-\subsection{Models}
+\subsubsection{Model comparison}
+\label{sec:model_comparison}
 
-The primary model for this analysis, RDG, is a constrained, L2-regularized linear regressor.
-Alternatives to this choice we explore include constrained and unconstrained gradient-boosted machine (GBM) models fit with the LightGBM \cite{LightGBM} algorithm as well as unconstrained L2-regularized linear regressors (i.e.
-RDG with no constraints).
-The performance of these alternatives, as quantified by relative risk estimates among top predictions, is illustrated in Figure~\ref{fig:performance_metric_mean_across_ta}.
+Figure~\ref{fig:performance_metric_mean_across_ta} presents average performance across therapeutic areas for select combinations of learning algorithm, constraint type and feature group described in Section~\ref{sec:models}.
+Several key findings illustrated in this figure are:
 
@@
   \caption{
-    \textbf{Performance of ...}.
+    \textbf{Performance across model algorithms and feature ablation groups}.
+    Average precision (AP) and receiver operating characteristic (ROC) scores with relative risk (RR) at ranking cutoffs denoted by \colorbox{Gainsboro}{RR@N}. 
+    (a) Performance across constrained RDG models using different groups of features as described in Section~\ref{sec:models}
+    (b) Performance across constrained (\colorbox{Gainsboro}{[+]}) and unconstrained (\colorbox{Gainsboro}{[+/-]}) linear and gradient-boosted models using the core feature set.
   }
@@
 
-We conclude from these results that 1) linear models outperform, or are at least not worse, than tree models when time is not used as a predictor and 2) non-negativity constraints in linear models result in greater performance.
-This is consistent with our expectation that all Open Targets scores should have monotonically increasing effects on the likelihood of success, since they were designed with this intent.
+\begin{enumerate}
+  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
+  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
+  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
+  \item Linear models outperform gradient-boosting models by nearly all measures
+  \item Constrained linear models outperform unconstrained linear models by nearly all measures
+\end{enumerate}
 
-\subsection{Sensitivity}
-
-In order to validate the stability of our findings in Section~\ref{sec:results_performance}, we repeat this analysis across 18 different configurations listed in Supplementary~Table~\ref{tab:sensitivity_configurations}.
-This includes 3 separate versions of Open Targets, 3 choices for the year defining the split between training and evaluation data and 2 choices for the length of the minimum advancement window (in years).
-
-We find that the mean RR values from the RDG model consistently exceed the OTS model in all configurations among the very highest rankings (N=10) and also exceed the OTS model in all configurations except for 1 for N between 20 and 60.
-This data is shown in Supplementary~Figure~\ref{fig:sensitivity_relative_risk}.
-The significance of these differences drops notably after N=40, which can be seen in the distribution of p-values from a Wilcoxon signed-rank test shown in Supplementary~Figure~\ref{fig:sensitivity_p_values}.
-
-\subsection{Predictions}
-
-Top predictions from the RDG model over the evaluation dataset are shown in Supplementary~Figure~\ref{fig:top_evaluation_predictions}.
-This figure demonstrates that there are a significant number of TD pairs with genetic support from many sources and that the combined effects of non-genetic features also substantially influence these rankings.
+We conclude from these results that constrained linear models are an optimal choice for this problem due both to their greater performance and interpretability.
+This interpretability is illustrated more in Section~\ref{sec:effects} and owed much to the effort Open Targets has already undertaken to construct evidence scores such that they can be assumed to have a monotonically increasing effect on the likelihood that a causal relationship exists between a target and a disease.
 
 \subsection{Effects}
+\label{sec:effects}
 
@@
 By comparison, scores for literature associations are typically far lower, even when limited only to cases where they exist, with a median value of 0.12 (mean=.23) in the evaluation data.
+This is why the \colorbox{Gainsboro}{europepmc} feature has a relatively large associated coefficient and a much smaller average effect on predictions.
 
@@
 In other words, we do not treat genetic evidence as longitudinal features like all others associated with TD pairs.
-A breakdown of which features are treated in which manner is provided in Supplementary~Table~\ref{tab:features}.
+A breakdown of which features are treated in which manner is provided in Table~\ref{tab:features}.
 
@@
   \item Discuss how typical classifier metrics are inadequate for this problem
-  \item Talk about how the inclusion of time greatly improves performance and suggests it may be possible to fill in this proxy with predictors for more specific early indicators of other reasons trials fail
+  \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
   \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
@@
   \item Describe how therapeutic areas were selected based on having at least 100 TD pairs with target-disease-specific evidence of any kind and with explicit omissions: ("biological\_process", "pregnancy or perinatal disease", "injury, poisoning or other complication", "pregnancy or perinatal disease", "medical procedure", "infectious disease", "animal disease")
+  \item RDG models are fit using the \colorbox{Gainsboro}{Ridge} implementation from scikit-learn and GBM models are fit using the LightGBM \cite{LightGBM} algorithm
   \item Training/evaluation results are limited by temporalization while the present-day predictions are not
@@
 
-\section{Supplementary Material}
+\section{Appendix}
 
@@
 
-\input{assets/tractability_bucket_table.tex}
+\begin{figure}[!htb]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{dataset_statistics.png}
+  \caption{
+    \textbf{Training and evaluation dataset summary statistics}.
+  }
+  \label{fig:dataset_statistics}
+\end{figure}
 
-\input{assets/sensitivity_configurations.tex}
+\section{Supplementary Material}
 
-\begin{figure}
+\subsection{Performance}
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
+  \caption{
+    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
+    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} indicate performance for the \colorbox{Gainsboro}{N} top TD pairs.
+The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years.
+The \colorbox{Gainsboro}{omim}, \colorbox{Gainsboro}{eva}, and \colorbox{Gainsboro}{ot\_genetics\_portal} features correspond to the OMIM, EVA and OTG baselines of Figure~\ref{fig:relative_risk_by_limit}, respectively.
+All other features are assessed based on their existence.
+The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
+  }
+  \label{fig:relative_risk_core_features}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}
+  \caption{
+    \textbf{Relative risk scores for target/disease features}.
+    The features ending with \colorbox{Gainsboro}{qtl\_Q} denote binary indicators constructed from cases where the feature meets or exceeds quantile \colorbox{Gainsboro}{Q} of its distribution.
+The features ending with \colorbox{Gainsboro}{gte\_X} denote indicators for when the feature meets or exceeds a specific value \colorbox{Gainsboro}{X}.
+  }
+  \label{fig:relative_risk_static_features}
+\end{figure}
+
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
+  \caption{
+    \textbf{Relative risk distributions across select therapeutic areas}.
+    P-values are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR averages across therapeutic areas exceed OTS averages.
+  }
+  \label{fig:relative_risk_dist_across_ta}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
+  \caption{
+    \textbf{Performance by algorithm, constraint type and feature grouping}.  
+  }
+  \label{fig:relative_risk_model_features}
+\end{figure}
+
+\subsection{Predictions}
+
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{dataset_statistics.png}
-  \caption{
-    \textbf{Training and evaluation dataset summary statistics}.
-  }
-  \label{fig:dataset_statistics}
-\end{figure}
+\subsection{Opportunities}
 
-\begin{figure}
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
-  \caption{
-    \textbf{Relative risk distributions across select therapeutic areas}.
-    P-values are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR averages across therapeutic areas exceed OTS averages.
-  }
-  \label{fig:relative_risk_dist_across_ta}
-\end{figure}
+\input{assets/tractability_bucket_table.tex}
 
-\begin{figure}
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}
-  \caption{
-    \textbf{Relative risk scores for target/disease features}.
-    The features ending with \colorbox{Gainsboro}{qtl\_Q} denote binary indicators constructed from cases where the feature meets or exceeds quantile \colorbox{Gainsboro}{Q} of its distribution.
-The features ending with \colorbox{Gainsboro}{gte\_X} denote indicators for when the feature meets or exceeds a specific value \colorbox{Gainsboro}{X}.
-  }
-  \label{fig:relative_risk_static_features}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
-  \caption{
-    \textbf{Performance by algorithm, constraint type and feature grouping}.  
-  }
-  \label{fig:relative_risk_model_features}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}
+\subsection{Sensitivity}
+
+In order to validate the stability of our findings in Section~\ref{sec:results_performance}, we repeat this analysis across 18 different configurations listed in Supplementary~Table~\ref{tab:sensitivity_configurations}.
+This includes 3 separate versions of Open Targets, 3 choices for the year defining the split between training and evaluation data and 2 choices for the length of the minimum advancement window (in years).
+
+We find that the mean RR values from the RDG model consistently exceed the OTS model in all configurations among the very highest rankings (N=10) and also exceed the OTS model in all configurations except for 1 for N between 20 and 60.
+This data is shown in Supplementary~Figure~\ref{fig:sensitivity_relative_risk}.
+The significance of these differences drops notably after N=40, which can be seen in the distribution of p-values from a Wilcoxon signed-rank test shown in Supplementary~Figure~\ref{fig:sensitivity_p_values}.
+
+\input{assets/sensitivity_configurations.tex}
+
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering

### 2024-04-01 analysis.py

@@ -1416,6 +1416,11 @@ def get_primary_models() -> PandasDataFrame:
                 "display_color": "#7f7f7f",  # grey
             },
             {
+                "model": "rdg__no_tgc__positive",
+                "model_slug": "RDG-X",
+                "display_color": "#7f7f7f",  # grey
+            },
+            {
                 "model": "gbm__all__unconstrained",
                 "model_slug": "GBM-T",
                 "display_color": "#ffffff",  # white

### 2024-04-06 paper/paper.tex
Add feature presence figure

@@
 
+\subsubsection{Metrics}
+
+The primary performance metric used in this study is relative risk (RR).
+One reason we use this metric is to maintain consistency similar studies like \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765}.
+
+as our primary performance indicator along with secondary measures of performance in more familiar classifier metrics like Receiver Operating Characteristic (ROC) and Average Precision (PR).
+as well as all others where not specified otherwise is defined as:
+
+\begin{equation}
+  \frac{P(advancement | rank >= N)}{P(advancement | rank < N)}.
+\end{equation}
+
 \subsubsection{Open Targets comparison}
@@
 Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study, RDG, ranks TD pairs by comparison to a composite score from Open Targets, OTS.
-The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:
+This comparison highlights relative risk (RR) as our primary performance indicator along with secondary measures of performance in more familiar classifier metrics like Receiver Operating Characteristic (ROC) and Average Precision (PR).
+as well as all others where not specified otherwise is defined as:
 
@@
 
-\input{assets/features_table.tex}
+\input{assets/feature_table.tex}
 
@@
 \section{Supplementary Material}
+
+\subsection{Features}
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \captionsetup[subfigure]{labelformat=empty}
+  \subfloat[\centering]{
+    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
+      \put(102, 20) {(a)}
+    \end{overpic}
+  }
+  \qquad
+  \subfloat[\centering]{
+    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
+      \put(102, 20) {(b)}
+    \end{overpic}
+  }
+  \caption{
+    \textbf{Evaluation dataset feature presence}.
+    (a) Feature presence drawn to scale. 
+    (b) Feature presence visually rescaled to fit counts. 
+  }
+  \label{fig:feature_presence}
+\end{figure}
 

### 2024-04-06 analysis.py

@@ -988,6 +988,53 @@ def get_feature_statistics(
     return feature_statistics
 
 
+def get_feature_presence(features: PandasDataFrame) -> PandasDataFrame:
+    def get_feature_group_label(fg: str) -> str:
+        entity, group = fg.split("__")
+        if entity == "target_disease":
+            entity = "pair"
+        if "phase_2" in group:
+            group = "phase 2+ trials"
+        elif group == "outcome":
+            group = "advanced beyond phase 2"
+        elif group == "no_data":
+            group = "no evidence"
+        else:
+            group = f"{group} evidence"
+        return f"{entity} has {group.replace('_', ' ')}"
+
+    # fmt: off
+    return (
+        features
+        .assign(target_disease_id=lambda df: df[['target_id', 'disease_id']].apply(tuple, axis=1))
+        .set_index('target_disease_id')
+        .assign(
+            target__phase_2=lambda df: df["target__clinical__phase_max__reached"] > 2,
+            disease__phase_2=lambda df: df["disease__clinical__phase_max__reached"] > 2,
+        )
+        .filter(regex='^target__phase|^disease__phase|^target_disease__(?!clinical|time)')
+        .assign(target_disease__no_data=lambda df: (
+            df.filter(regex='^target_disease__(?!outcome)')
+            .pipe(lambda df: df.fillna(0) <= 0)
+            .all(axis=1)
+        ))
+        .pipe(lambda df: df.where(df > 0))
+        .pipe(apply, lambda df: logger.info(
+            f"The following features will be used to calculate presence: {df.columns.tolist()}"
+        ))
+        .rename_axis("feature", axis="columns")
+        .stack().rename("value").reset_index()
+        .assign(feature_group=lambda df: df['feature'].str.split('__').str[:2].str.join("__"))
+        .groupby("feature_group")["target_disease_id"].unique()
+        .rename("pairs").reset_index()
+        .assign(n_pairs=lambda df: df['pairs'].apply(len))
+        .assign(feature_group_label=lambda df: df['feature_group'].map(get_feature_group_label))
+        .sort_values("n_pairs", ascending=True)
+        [["feature_group", "feature_group_label", "n_pairs", "pairs"]]
+    )
+    # fmt: on
+
+
 def get_feature_relative_risk(
     ds: Dataset,
     model_names: list[str] | None = None,
@@ -2385,6 +2432,10 @@ def get_assets_directory() -> Path:
     if not path.exists():
[…]

### 2024-04-07 paper/paper.tex
Add section on performance metrics

--- 2024-04-01 paper/paper.tex
@@
 
-It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:36963162}.
+It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:37803084,PMID:36963162}, and to a lesser extent, that the same may also be true for single-cell transcriptomic evidence \cite{Dann2024.04.04.24305313}.
 This information has been used to devise many target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}.
@@
 
+\subsection{Metrics}
+\label{sec:metrics}
+
+The primary performance metric used in this study is relative risk (RR).
+This metric is commonly used to assess univariate measures of genetic support \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765} and can be more intuitively understood, in the context of this study, as the probability that a TD pair among the top $N$ TD pairs as ranked by a particular method will advance beyond phase 2 trials divided by that same probability of advancement among TD pairs with a rank greater than $N$.
+This provides a means to compare multivariate, model-based methods to univariate methods on a common scale.
+More specifically, any RR metric reported for a model among top $N$ rankings is defined as:
+
+\begin{equation}
+  \frac{P(advancement | rank >= N)}{P(advancement | rank < N)}
+\end{equation}
+
+and RR metrics reported for univariate methods based on Open Targets scores for a single type of evidence, where not stated otherwise, are defined as:
+
+\begin{equation}
+  \frac{P(advancement | score > 0)}{P(advancement | score = 0)}.
+\end{equation}
+
+The use of such a metric is essential for properly assessing performance in this forecasting problem.
+While we also report more common measures of classifier performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), neither of these adequately capture behavior in the upper extremes of rankings due the sparsity with which TD pair evidence is present for pairs that have ever entered phase 2 trials.
+This sparsity is further exacerbated in this study by the constraint that most of that evidence must have existed \textit{before} such trials began.
+Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=9010), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
+These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases.
+Specifically, 8,425 (93.5\%) have a target and 6,855 (76.1\%) have a disease that had already been in phase 2 or later trials previously.
+Figure~\ref{fig:feature_presence} also demonstrates that a substantial fraction of these pairs that ultimately advance beyond phase 2 trials either have no direct evidence or have only prior clinical evidence for the target or disease alone, which means a comprehensive measure of classifier performance (e.g.
+ROC) is far more likely to reflect the extent to which disease-only historical, clinical information or target-only information -- including other attributes like conservation, essentiality and tissue expression -- can predict clinical success.
+Again, this is not our primary focus as we want to evaluate the maximum achievable performance in this forecasting problem, and we assert that this is best accomplished when one or more lines of target-disease-specific evidence are present.
+
+Reasonable alternative choices for this primary metric include those more common in information extraction literature or other machine learning studies with a focus on ranking rather than classification, such as mean reciprocal rank (MRR), precision at $k$ (P@k) and normalized discounted cumulative gain (NDCG) \cite{hoyt2022unified,moffat2022batch}.
+Precision at $k$ is the most similar among these to relative risk at $k$ since it is equivalent to the numerator in the relative risk calculation.
+We use relative risk instead because it is more intuitive than most ranking metrics, has well established analytical solutions for confidence intervals \cite{Katz1978-mo} and is consistent with prior work in this field.
+
+Lastly, we emphasize that the interpretation of "risk" for the relative risk metric is to be inverted in this context.
+A higher "risk" in this study actually corresponds to a greater probability of success.
+The name "Relative Success" is used for this metric instead in \cite{Minikel2023.06.23.23291765} even though it has the same underlying definition.
+We choose not to use this label because we also present generic performance measures like ROC and AP, thereby prioritizing consistency with a domain-independent nomenclature where possible.
+
 \subsection{Performance}
@@
 Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study, RDG, ranks TD pairs by comparison to a composite score from Open Targets, OTS.
-The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:
-
-\begin{equation}
-  \frac{P(advancement | rank >= N)}{P(advancement | rank < N)}.
-\end{equation}
-
+This comparison highlights relative risk (RR) as our primary performance indicator along with secondary measures of performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), as discussed more in Section~\ref{sec:metrics}.
 The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it uses time since the phase 2 transition as a predictive factor in addition to all others.
-We observe that the use of this information almost universally improves standard performance metrics like receiver operating characteristic (ROC) and average precision (AP), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed.
+We observe that the use of this information greatly improves standard performance metrics like receiver operating characteristic (ROC) and average precision (AP), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed.
 In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support.
@@
   \item Discuss the possibility to do prospective evaluation with OT snapshots
-  \item Discuss how typical classifier metrics are inadequate for this problem
+  \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
   \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
@@
 
-\input{assets/features_table.tex}
+\subsection{Features}
+
+\input{assets/feature_table.tex}
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \captionsetup[subfigure]{labelformat=empty}
+  \subfloat[\centering]{
+    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
+      \put(102, 20) {(a)}
+    \end{overpic}
+  }
+  \qquad
+  \subfloat[\centering]{
+    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
+      \put(102, 20) {(b)}
+    \end{overpic}
+  }
+  \caption{
+    \textbf{Evaluation dataset feature presence}.
+    (a) Feature presence drawn to scale. 
+    (b) Feature presence visually rescaled to fit labels and combinatorial patterns. 
+  }
+  \label{fig:feature_presence}
+\end{figure}
 

### 2024-04-08 paper/paper.tex
Finish integration of feature presence exposition

@@
 \subsection{Features}
+\label{sec:features}
 
@@
 Examples of this include target-specific tissue expression specificity scores computed from Human Protein Atlas \cite{PMID:25613900} and LOEUF \cite{PMID:32461654} scores from gnomAD.
-Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OT23.12release} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features.
+Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OT23.12release} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (described in Section~\ref{sec:methods}).
+
+\begin{figure}[!htb]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \captionsetup[subfigure]{labelformat=empty}
+  \subfloat[\centering]{
+    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
+      \put(102, 20) {(a)}
+    \end{overpic}
+  }
+  \qquad
+  \subfloat[\centering]{
+    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
+      \put(102, 20) {(b)}
+    \end{overpic}
+  }
+  \caption{
+    \textbf{Evaluation dataset feature presence}.
+    (a) Presence and co-occurrence of various feature groupings or individual features indicated by overlapping sets of TD pairs, where the size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin).
+The top margin shows the number of overlapping features for a combination of features and the bottom margin shows the number of TD pairs that are a part of that same combination.
+    (b) Same as (a) without truly proportional bar sizes that are not scaled down beyond a limit necessary to fit all margin labels and combinatorial groupings.
+Features that are not specific to TD pairs are also omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
+  }
+  \label{fig:feature_presence}
+\end{figure}
+
+The combination of target, disease and target-disease features creates sparsity patterns that are important to understand before interpreting models built from it.
+Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=9010), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
+These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases.
+Specifically, 8,425 (94\%) have a target and 6,855 (76\%) have a disease that had already been in phase 2 or later trials previously.
+This is consistent with herding effects observed in recent drug development pipelines \cite{PMID:37117303} over the same time period (2016-2022) and underscores the prognostic value such information may have as it is becoming more and more common and clearly confers lower clinical risk for new drug programs.
+We also observe that, on top of the clear theoretical, causal relationship between prior target and disease human clinical validation and the likely success of programs for new combinations of such targets and diseases, these features have strong, univariate predictive effects in the evaluation dataset of our study.
+This is illustrated in Supplementary~Figure~\ref{fig:relative_risk_static_features}, which shows the relative risk of advancement for these features capturing the highest clinical stage previously reached by a target or disease.
+
+Taken together, the paucity of TD pair evidence and the abundance of prior target or disease clinical validation should be considered carefully when interpreting predictive performance.
+TD pairs predicted to be highly likely to advance clinically despite a lack of target-disease-specific evidence are entirely plausible, but the value of such a prediction is dependent upon the application.
+We choose to minimize the influence of these cases in our study by focusing on rankings within therapeutic areas that do not extend beyond the number of TD pairs with direct evidence of some kind, as discussed in Section~\ref{sec:methods}.
+This choice is also reflected in our primary performance metrics, as discussed in Section~\ref{sec:metrics}.
 
@@
 This sparsity is further exacerbated in this study by the constraint that most of that evidence must have existed \textit{before} such trials began.
-Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=9010), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
-These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases.
-Specifically, 8,425 (93.5\%) have a target and 6,855 (76.1\%) have a disease that had already been in phase 2 or later trials previously.
-Figure~\ref{fig:feature_presence} also demonstrates that a substantial fraction of these pairs that ultimately advance beyond phase 2 trials either have no direct evidence or have only prior clinical evidence for the target or disease alone, which means a comprehensive measure of classifier performance (e.g.
+For more details on the extent of this sparsity, see Section~\ref{sec:features}.
+The figure presented there, Figure~\ref{fig:feature_presence}, also demonstrates that a substantial fraction (78\%) of TD pairs that ultimately advance beyond phase 2 trials have targets and/or diseases with prior clinical validation despite no direct evidence linking the pairs themselves, which means a comprehensive measure of classifier performance (e.g.
 ROC) is far more likely to reflect the extent to which disease-only historical, clinical information or target-only information -- including other attributes like conservation, essentiality and tissue expression -- can predict clinical success.
@@
 The name "Relative Success" is used for this metric instead in \cite{Minikel2023.06.23.23291765} even though it has the same underlying definition.
-We choose not to use this label because we also present generic performance measures like ROC and AP, thereby prioritizing consistency with a domain-independent nomenclature where possible.
+We choose not to use this label because we also present generic performance measures like ROC and AP, thereby prioritizing consistency with a domain-independent nomenclature.
 
@@
   \item Discuss the possibility to do prospective evaluation with OT snapshots
+  \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
   \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
@@
   \item RDG models are fit using the \colorbox{Gainsboro}{Ridge} implementation from scikit-learn and GBM models are fit using the LightGBM \cite{LightGBM} algorithm
+  \item Cite supervenn
   \item Training/evaluation results are limited by temporalization while the present-day predictions are not
@@
 \input{assets/feature_table.tex}
-
-\begin{figure}[H]
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \captionsetup[subfigure]{labelformat=empty}
-  \subfloat[\centering]{
-    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
-      \put(102, 20) {(a)}
-    \end{overpic}
-  }
-  \qquad
-  \subfloat[\centering]{
-    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
-      \put(102, 20) {(b)}
-    \end{overpic}
-  }
-  \caption{
-    \textbf{Evaluation dataset feature presence}.
-    (a) Feature presence drawn to scale. 
-    (b) Feature presence visually rescaled to fit labels and combinatorial patterns. 
-  }
-  \label{fig:feature_presence}
-\end{figure}
 

### 2024-04-08 analysis.py

@@ -994,7 +994,7 @@ def get_feature_presence(features: PandasDataFrame) -> PandasDataFrame:
         if entity == "target_disease":
             entity = "pair"
         if "phase_2" in group:
-            group = "phase 2+ trials"
+            group = "prior phase 2+ trials"
         elif group == "outcome":
             group = "advanced beyond phase 2"
         elif group == "no_data":

### 2024-04-08 paper/paper.tex
Abstract re-write

@@
 \begin{abstract}
-This study examines the extent to which the outcomes of clinical trials can be predicted based on longitudinal properties of drug targets and diseases alone.
-We find that this is possible by comparing the historical performance of model-based target-disease pair prioritization methods to common baselines.
-Our primary objective is to demonstrate that statistical learning can effectively optimize such methods with no loss of interpretability.
-For example, non-negative linear models can produce simple weighting schemes across various types of human, animal and cell model evidence (for targets, diseases and pairings of the two) to identify target-disease pairs that advance beyond phase 2 trials with an average relative risk that is 2x higher than Open Targets composite scores.
-Other key characteristics of this study include: 1) a comprehensive longitudinal treatment of evidence as well as how it relates to leakage and reverse causality in biomedical research, 2) trial and/or drug details are not used in order to enable ranking for undeveloped targets/diseases, 3) analysis of the space of currently undeveloped, tractable targets with the highest likelihood of clinical success, 3) no data is used outside of Open Targets to ease reproduction and/or deployment, and 4) our method requires no expert knowledge and can easily support the inclusion of more lines of evidence over time, making it easy to operationalize.
+
+Selecting optimal drug targets for diseases represents a critical juncture in the creation of new medicines.
+In this study, we present an automated framework for selecting such targets through a synthesis of human, animal, and cell model evidence that features simple, highly interpretable statistical models.
+These models are optimized based on the historical clinical success of target-disease pairings, use only 33 predictors derived directly from individual lines of evidence in Open Targets and can predict clinical advancement for those pairings with a success rate that is 1.5-2x higher than heuristic, composite scores derived from the same evidence as well as common measures of genetic support that have already been established -- and observed in this study -- to confer a 2x higher likelihood of success.
+We find that this cumulative 3-4x advantage is maximized when also utilizing prior human clinical validation specific to targets and diseases independent of pairings, or other static properties of targets like conservation, expression specificity and gene essentiality.
+These predictors are rarely absent, which contrasts highly with the evidence for target-disease pairs that rarely exists among those that have ever reached clinical trials.
+The extent and non-random nature with which this absence manifests pose key challenges in the selection of performance measures and predictor derivation.
+In particular, we address the possibility for reverse causality between clinical outcomes and evidence discovery through the use of longitudinal predictors.
+This novel aspect of our work also enables us to directly evaluate which types of evidence are most likely to leak information about clinical outcomes when not temporalized.
+By formulating target selection as a machine learning problem in this manner, and after accounting for these unique challenges, this framework can be easily deployed to prioritize disease targets yet to undergo clinical development.
+We explore some of these present-day targets, with a focus on those that are tractable, and offer all code and data used in this analysis as a proof-of-concept for our more comprehensive, proprietary ranking platform.
 % Note: OT multiplier (2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10

### 2024-04-08 paper/paper.tex
Refactor introduction

@@
 It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:37803084,PMID:36963162}, and to a lesser extent, that the same may also be true for single-cell transcriptomic evidence \cite{Dann2024.04.04.24305313}.
-This information has been used to devise many target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}.
-It is also possible to expand the breadth of this genetic support to more targets/diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}.
+This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}.
+It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}.
 To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates.
-This is not a focus of this work as we aim, instead, to establish a framework for identifying targets/diseases with the very highest possible likelihood of success first.
-This is accomplished by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in a simple statistical modeling framework.
-It can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret.
-We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence from many sources, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.
+This is not a focus of this work as we aim, instead, to establish a framework for identifying target-disease (TD) pairs with the very highest possible likelihood of success.
+We accomplish this by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in statistical models that are trained on past clinical outcomes, evaluated primarily on their performance at upper extremes of predictions and then deployed on undeveloped TD pairs to establish an evidence-based ranking pipeline.
+This method can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret.
+We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence, spanning hundreds or even thousands of putative orthogonal, causal predictors, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.
+This is the basis for our internal ranking platform, and such operational advantages have made it possible to reach or exceed the inclusivity of highly integrative methods without sacrificing interpretability, modularity and ease of deployment.
 
 A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine.
-This is vital because any method that optimizes for likely future clinical success based on historical clinical success may easily be biased by the non-random nature with which evidence is absent otherwise.
-We use only "temporalized" evidence, i.e.
-evidence for which timing of its emergence can be determined, and outcomes when training and evaluating our methods before ultimately applying them to present-day evidence with no restrictions on timing.
-We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.
+This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent otherwise.
+We use longitudinal evidence, i.e.
+evidence for which timing of its emergence can be determined, where appropriate when training and evaluating our methods before ultimately applying them to present-day evidence that is not temporalized.
+We discuss motivations, prior research and our own analysis on how important this problem is for each type of evidence in Section~\ref{sec:results_inflation}.
 
-Addressing such problems is common, but not ubiquitous, in studies that attempt to predict clinical trial outcomes using temporalized predictors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}.
-The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets/diseases, trial sponsor track records, eventual patient enrollment, etc.
+While the use of longitudinal evidence is very rare in studies on target prioritization algorithms, it is not at nearly as rare in studies that attempt to predict clinical trial outcomes using a variety of biological evidence sources, clinical statistics and other organizational or behavioral factors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}.
+The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets or diseases, trial sponsor track records, eventual patient enrollment, etc.
 constitute clear information leaks otherwise.
-This is discussed more in \cite{PMID:37483175} which notes several studies that do not account for this problem, and presents "quasi-prospective" as well as true "prospective" results.
-The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until years later.
-Nomenclature for these formulations is conflicting though, where this definition of a "quasi-prospective" design is deemed entirely prospective in some cases, e.g.
+This is discussed in some depth in \cite{PMID:37483175} which notes several studies that do not account for this problem before drawing a clear distinction between quasi-prospective and prospective problem formulations.
+The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until sufficient time has elapsed for more outcomes to occur.
+Nomenclature for these formulations is conflicting though, where this definition of a quasi-prospective formulation is deemed entirely prospective in some cases, e.g.
 \cite{PMID:37225853}.
-We will refer to our design in this study as quasi-prospective since this definition is the best fit.
+To avoid any potential confusion, our formulation in this study is quasi-prospective.
+The distinct advantages and disadvantages of each formulation provide a complementary risk/reward trade-off that we capitalize on in our internal platform, and we assert that the distinction between them is important.
+Interestingly, the existence of Open Targets snapshots dating back to at least 2016 would enable the extension of this work to a prospective formulation as well.
+This is discussed more in Section~\ref{sec:discussion}.
 
-The prior works discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with clinical trial success and 2) clinical trial outcome prediction models.
-Both are measured against the same outcomes and an important distinction between them lies in how the former methods are {\bf not} directly optimized for those outcomes while the latter methods are.
-In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial.
+The preceding works discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success and 2) clinical trial outcome prediction models.
+Both are measured against the same outcomes and an important distinction between them lies in how the prioritization methods are {\bf not} directly optimized for those outcomes while the outcome prediction methods are.
+In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that could be present well in advance of any drug program or individual trial.
 We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS.
 Finally, we examine how many present-day target-disease pairs are undeveloped (i.e.
-have never been in clinical trials), have a tractable target and are likely to see success rates matching or exceeding those calibrated benchmarks.
+have never been in clinical trials), have a tractable target and are likely to see success rates matching or exceeding those benchmarks.
 
@@
 
-In order to model clinical advancement for target-disease (TD) pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with any one TD pair as indicated by the presence of a later-stage trial.
+In order to model clinical advancement for target-disease pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with any one TD pair as indicated by the presence of a later-stage trial.
 All results to follow consider only advancement beyond phase 2 due to limitations described in Section~\ref{sec:discussion}.

### 2024-04-09 paper/paper.tex
Start discussion section

@@
 \begin{abstract}
-
-Selecting optimal drug targets for diseases represents a critical juncture in the creation of new medicines.
-In this study, we present an automated framework for selecting such targets through a synthesis of human, animal, and cell model evidence that features simple, highly interpretable statistical models.
-These models are optimized based on the historical clinical success of target-disease pairings, use only 33 predictors derived directly from individual lines of evidence in Open Targets and can predict clinical advancement for those pairings with a success rate that is 1.5-2x higher than heuristic, composite scores derived from the same evidence as well as common measures of genetic support that have already been established -- and observed in this study -- to confer a 2x higher likelihood of success.
-We find that this cumulative 3-4x advantage is maximized when also utilizing prior human clinical validation specific to targets and diseases independent of pairings, or other static properties of targets like conservation, expression specificity and gene essentiality.
-These predictors are rarely absent, which contrasts highly with the evidence for target-disease pairs that rarely exists among those that have ever reached clinical trials.
-The extent and non-random nature with which this absence manifests pose key challenges in the selection of performance measures and predictor derivation.
-In particular, we address the possibility for reverse causality between clinical outcomes and evidence discovery through the use of longitudinal predictors.
-This novel aspect of our work also enables us to directly evaluate which types of evidence are most likely to leak information about clinical outcomes when not temporalized.
-By formulating target selection as a machine learning problem in this manner, and after accounting for these unique challenges, this framework can be easily deployed to prioritize disease targets yet to undergo clinical development.
-We explore some of these present-day targets, with a focus on those that are tractable, and offer all code and data used in this analysis as a proof-of-concept for our more comprehensive, proprietary ranking platform.
-% Note: OT multiplier (2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10
+Choosing which drug targets to pursue for a given disease is one of the most impactful decisions made in the global development of new medicines.
+This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases.
+We demonstrate a novel statistical learning framework utilizing both constrained and unconstrained linear and tree models, that can optimize predictions of a target-disease pair's likelihood of advancing beyond phase 2 clinical trials, outperforming existing baselines, including Open Targets composite scores, by 2x or greater, while preserving interpretability and counteracting common forms of bias.
+Utilizing a subset of our biomedical evidence base, non-negative linear models can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify target-disease pairs with high probabilities of advancing beyond phase 2 trials, and uncover a range of previously undeveloped target-disease pairs poised for clinical success.
+In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different type of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success.
+To ease reproduction and deployment, none of our proprietary data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
+% Note: OT multiplier (1.5-2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10 and mean@20
 \end{abstract}
@@
 
+We have focused primarily on performance among the top 1\% of TD pairs throughout the majority of this study.
+Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}.
+It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance for the remaining 99\%.
+We offer some evidence of this here through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}).
+
 \begin{itemize}
+  \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
+  \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
+  \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
   \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
@@
   \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
-  \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
-  \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
-  \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
   \item "There is evidence that a 9.6\% vs.

### 2024-04-10 paper/paper.tex
Draft first half of discussion section

@@
 
+%% Variables
+\def\evaluationDatasetPairCount{9010}
+\def\topRankingPct{2}
+\def\bottomRankingPct{98}
+
 \begin{abstract}
@@
 This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases.
-We demonstrate a novel statistical learning framework utilizing both constrained and unconstrained linear and tree models, that can optimize predictions of a target-disease pair's likelihood of advancing beyond phase 2 clinical trials, outperforming existing baselines, including Open Targets composite scores, by 2x or greater, while preserving interpretability and counteracting common forms of bias.
+We demonstrate a novel statistical learning framework utilizing both constrained and unconstrained linear and tree models, that can predict clinical advancement for the top \topRankingPct\% of target-disease pairs 1.5-2x more effectively than Open Targets composite scores as well as common measures of genetic support that have already been established -- and observed in this study -- to confer a 2x higher likelihood of success (a 3-4x cumulative improvement).
 Utilizing a subset of our biomedical evidence base, non-negative linear models can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify target-disease pairs with high probabilities of advancing beyond phase 2 trials, and uncover a range of previously undeveloped target-disease pairs poised for clinical success.
 In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different type of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success.
-To ease reproduction and deployment, none of our proprietary data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
-% Note: OT multiplier (1.5-2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10 and mean@20
+To ease reproduction and deployment, no data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
+% Note: OT multipliers come from Sensitivity / Relative risk ratio distribution across configurations / mean@10 and mean@20
 \end{abstract}
@@
 The combination of target, disease and target-disease features creates sparsity patterns that are important to understand before interpreting models built from it.
-Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=9010), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
+Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=\evaluationDatasetPairCount), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
 These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases.
@@
       \put(50, -3) {(a)}
+      % Performance multipliers
+      \put(10.5,28.5){\line(0,1){27.5}} % Vertical line
+      \put(10.5,28.5){\line(1,0){.5}} % Bottom bar
+      \put(10.5,56){\line(1,0){.5}} % Top bar
+      \put(6,42.5){2x}
+      \put(30.5,22.25){\line(0,1){6.5}}
+      \put(23,25){1.5x}
+      % Ranking percentages
+      \put(19,9){\line(0,1){4}}
+      \put(16,14){Top 1\%}
+      \put(36.5,9){\line(0,1){4}}
+      \put(33.5,14){Top 2\%}
     \end{overpic}
@@
     See Supplementary~Figure~\ref{fig:relative_risk_by_ta} for raw data underlying (a).
+    Annotations indicate RR multiples between RDG and OTS for the top 10 and top 30 TD pairs as well as how many TD pairs comprise the top 1\% (N=18) and top 2\% (N=36) of rankings on average across therapeutic areas.
+All results shown are from the evaluation dataset.
+    % Ranking percentages are based on an average of 1796 TD pairs per therapeutic area
   }
@@
   \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_by_limit.pdf}
+  \begin{overpic}[width=1\textwidth]{relative_risk_by_limit.pdf}
+    % Ranking percentages
+    \put(23,8){\line(0,1){2}}
+    \put(20,11){Top 1\%}
+    \put(37,8){\line(0,1){2}}
+    \put(34,11){Top 2\%}
+  \end{overpic}
   \caption{
     \textbf{Performance compared to genetic support benchmarks}.
-The RR estimates for each benchmark are based on the presence of the corresponding support across all TD pairs, and the number of pairs for which is present is represented by the horizontal bars extending horizontally from the y-axis.
-The bounds around the RDG RR estimates correspond to a Katz 90\% confidence interval.
+The RR estimates for each benchmark annotated on the right are shown as horizontal dotted lines and are calculated across all TD pairs (N=\evaluationDatasetPairCount).
+The number of pairs with support from each benchmark is represented by the bars extending horizontally (to the right) from the y-axis.
+The RDG RR point estimate is shown in blue and bounds around that estimate correspond to a Katz 90\% confidence interval.
+Annotations on the x-axis indicate how many TD pairs comprise the top 1\% (N=90) and top 2\% (N=180) of rankings.
+All results shown are from the evaluation dataset.
   }
@@
 By comparison, scores for literature associations are typically far lower, even when limited only to cases where they exist, with a median value of 0.12 (mean=.23) in the evaluation data.
-This is why the \colorbox{Gainsboro}{europepmc} feature has a relatively large associated coefficient and a much smaller average effect on predictions.
+This is why the \colorbox{Gainsboro}{europepmc} feature has a relatively large associated coefficient, but a much smaller average effect across predictions.
 
@@
 
-We have focused primarily on performance among the top 1\% of TD pairs throughout the majority of this study.
+An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery.
+One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}.
+If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug.
+Pushing that even further to 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively.
+These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence, however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e.
+untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.
+
+% TODO: complete this section?
+Another way to quantify the impact that employing superior target prioritization strategies could have is to consider what would happen if this was done across the entire pharmaceutical development industry.
+Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year.
+This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ...
+
+Throughout the majority of this study, we focus primarily on performance among the top \topRankingPct\% of TD pairs.
 Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}.
-It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance for the remaining 99\%.
-We offer some evidence of this here through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}).
+It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance for the remaining \bottomRankingPct\%.
+We offer some evidence of this through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}).
+While time is a necessary but not sufficient condition for clinical advancement of a TD pair, given that a single trial can take years to complete, it also correlates with how many distinct trials, drugs, companies, sponsors, investigators, etc.
+attempt to validate any one pair.
+We posit that this is crucial because even if a druggable, mechanistic link exists between a target and a disease, many (or even most) trials testing that link will fail for reasons unrelated to efficacy or safety of a drug \cite{Razuvayevskaya2023.02.07.23285407}.
+Some of these reasons include 1) commercial factors like a lack of funding, pipeline reprioritization and competitive density, 2) administrative or logistical factors like a lack of enrollment, retention problems, poor trial design, drug supply chain shortages, epidemics (COVID-19) and 3) regulatory factors like legislative shifts or extensive regulatory approval delays.
+Many of these factors are fundamentally difficult or impossible to predict.
+Epidemic outbreaks, labor shortages leading to supply chain problems and regulatory shifts are examples of exogenous shocks for which accurate predictors are unlikely to exist.
+However, factors like funding, competition, enrollment and certain aspects of trial design have measurable predictors such as preceding venture capital and private equity investments, patent filings and revenue for related drugs, disease severity and prevalence, and the existence of biomarkers or other enabling factors for better trial designs (respectively).
+As a more detailed example, patient recruitment failures are the most common reason trials fail \cite{Razuvayevskaya2023.02.07.23285407} and recruitment statistics are highly predictive of eventual trial outcomes \cite{Lo2019Machine,PMID:34430930}.
+It stands to reason then that predicting trial enrollment failures well in advance of when a trial is even conceived, i.e.
+for a target-disease pair, should be possible because recruitment in trials is determined, in part, by how prevalent a disease is, how debilitating it is, what patient demographics it inflicts, whether existing treatments exist for it and how successful recruitment for past trials (if any exist) has been -- all of which are measurable.
+Signals like this could be used to further improve performance in the bottom \bottomRankingPct\% of TD pairs since they better capture causal factors in non-biological trial failures that time under development -- the only non-biological feature in this study -- does not.
+
+Another notable focus of this study is on evidence for TD pairs that can be \textbf{directly} attributed to them.
+This is a departure from earlier research in this space like \cite{Nelson2015-eg} that often includes measures of similarity between disease ontology terms to account for the fact that bridging disease nomenclatures used in clinical trial datasets with those used in databases maintaining genetic evidence is difficult.
+It is not uncommon for disease terms from either source to be sufficiently similar such that it is appropriate to consider them as equivalent.
+It can also be argued that similarity between disease terms offers an important dimension for expanding evidence in a biologically meaningful way, regardless of technical mapping issues.
+This is taken to further extremes in studies like \cite{PMID:33262371} that propagates evidence using both measures of target similarity (from protein interaction networks) and measures of disease similarity (from ontologies and literature co-occurrence).
+We believe such expansion methods are very compatible with our approach by either using them to derive new predictors or to propagate predictions from models like these across networks or knowledge graphs.
 
 \begin{itemize}
-  \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
-  \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
-  \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
+  \item Discuss the possibility to do prospective evaluation with OT snapshots
   \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
-  \item Discuss the possibility to do prospective evaluation with OT snapshots
   \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
-  \item "There is evidence that a 9.6\% vs.
-13.8\% success rate for drugs from phase 1 trials to approval may mean a \$480 million difference in the median research and development cost required to bring a new drug to the market (Wouters, McKee, and Luyten 2020)." \cite{PMID:34930919}
   \item From \cite{PMID:33262371}: "It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices.
 Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid."
-  \item Add select tractability and DepMap features?
 \end{itemize}

### 2024-04-11 paper/paper.tex
Add pagebreaks in supplements
Finish initial Methods draft

@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
 These are listed in Table~\ref{tab:features}.
-The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods that do not favor well-studied or drugged targets/diseases.
-Examples of this include target-specific tissue expression specificity scores computed from Human Protein Atlas \cite{PMID:25613900} and LOEUF \cite{PMID:32461654} scores from gnomAD.
-Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OT23.12release} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (described in Section~\ref{sec:methods}).
+The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods.
+This choice was motivated in part by empirical results presented in Section~\ref{sec:results_inflation} and assumptions described more in Section~\ref{sec:methods}.
+Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OTtargetPrioritisation} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (also described in Section~\ref{sec:methods}).
 
@@
     \textbf{Evaluation dataset feature presence}.
-    (a) Presence and co-occurrence of various feature groupings or individual features indicated by overlapping sets of TD pairs, where the size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin).
-The top margin shows the number of overlapping features for a combination of features and the bottom margin shows the number of TD pairs that are a part of that same combination.
-    (b) Same as (a) without truly proportional bar sizes that are not scaled down beyond a limit necessary to fit all margin labels and combinatorial groupings.
-Features that are not specific to TD pairs are also omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
+    (a) Presence and co-occurrence of select features for overlapping sets of TD pairs.
+The size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin).
+The top margin shows the number of features in any one feature combination and the bottom margin shows the number of TD pairs that are a part of that same combination.
+    (b) Same as (a) with proportional bar scaling that does not drop beyond a limit necessary to fit all margin labels and combinatorial groupings.
+Features that are not specific to TD pairs are omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
   }
@@
 The figure presented there, Figure~\ref{fig:feature_presence}, also demonstrates that a substantial fraction (78\%) of TD pairs that ultimately advance beyond phase 2 trials have targets and/or diseases with prior clinical validation despite no direct evidence linking the pairs themselves, which means a comprehensive measure of classifier performance (e.g.
-ROC) is far more likely to reflect the extent to which disease-only historical, clinical information or target-only information -- including other attributes like conservation, essentiality and tissue expression -- can predict clinical success.
+ROC) is far more likely to reflect the extent to which disease-only historical, clinical information or target-only information -- including other attributes like conservation and tissue expression -- can predict clinical success.
 Again, this is not our primary focus as we want to evaluate the maximum achievable performance in this forecasting problem, and we assert that this is best accomplished when one or more lines of target-disease-specific evidence are present.
@@
 
+We conclude that the RDG model outperforms OTS by a factor of at most 2 and least 1.5 for a significant number of TD pairs.
+A sensitivity analysis for this key result is presented in Supplementary~Section~\ref{sec:sensitivity}, where we find that these multipliers hold on average across multiple Open Targets releases and other important settings.
+
 \subsubsection{Genetic benchmark comparison}
@@
 One key objective of this study is to determine if any model, e.g.
-RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeding what is expected from any one source of genetic support alone.
-We find that this goal is met and exceeded by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support alone at an expected rate of advancement exceeding that of the single source (respectively).
+RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support.
+We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement.
 This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult.
@@
 The RR estimates for each benchmark annotated on the right are shown as horizontal dotted lines and are calculated across all TD pairs (N=\evaluationDatasetPairCount).
-The number of pairs with support from each benchmark is represented by the bars extending horizontally (to the right) from the y-axis.
-The RDG RR point estimate is shown in blue and bounds around that estimate correspond to a Katz 90\% confidence interval.
+The number of pairs with support from each benchmark is represented by the bars extending horizontally (left to right) from the y-axis.
+RDG RR point estimates are shown in blue and bounds around the estimates correspond to Katz 90\% confidence intervals.
 Annotations on the x-axis indicate how many TD pairs comprise the top 1\% (N=90) and top 2\% (N=180) of rankings.
@@
 Similar comparisons for target-specific and disease-specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}.
-These findings suggest that 1) genetic support for TD pairs is highly predictive but rare, 2) human clinical support for targets and diseases in isolation is also predictive while being more common and 3) constraint and expression specificity of targets exhibit modest but significant effects.
-While only human clinical evidence appears to have a prognostic value rivaling that of genetic support when considered on a univariate basis, the combined influence of non-genetic, non-clinical evidence is examined in Section~\ref{sec:model_comparison} where a model using only this information, RDG-X, still outperforms Open Targets composite scores by all measures.
 
@@
   \captionsetup{width=.9\linewidth}
-	\includegraphics[width=1\textwidth]{effect_sizes.pdf}
+  \begin{overpic}[width=1\textwidth]{effect_sizes.pdf}
+    \put(48, .5) {(a)}
+    \put(70, .5) {(b)}
+  \end{overpic}
   \caption{
     \textbf{RDG model feature effects}. 
-The \textbf{effect max} values are equivalent to RDG model coefficients for the corresponding feature while the \textbf{effect mean} values indicate average values of the product between the coefficient and a particular feature value, when that feature is present.
+    (a) Feature average effects calculated as the mean of the product between a coefficient and a particular feature value, when that feature is present and non-negative (i.e.
+when it has any influence).
+Only the evaluation dataset is used to compute this average.
+    (b) Feature coefficients, which are equivalent to the maximum effect that a feature can have due to all features being defined on the range [0, 1] or rescaled to it.
   }
@@
 
-A common method for identifying druggable opportunities within a specific disease context involves first ranking TD pairs according to some prioritization methodology followed by filtering or reprioritizing those ranks based on knowledge of target tractability \cite{PMID:28356508,PMID:35401535,PMID:31253980}.
+A common method for identifying druggable opportunities within a specific disease context involves first ranking TD pairs according to some prioritization methodology followed by filtering or re-prioritizing those ranks based on knowledge of target tractability \cite{PMID:28356508,PMID:35401535,PMID:31253980}.
 We use a similar approach to identify tractable targets associated with TD pairs that have yet to enter clinical trials.
@@
 Top antibody-enabled pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.
+
+We also note that these TD pairs are identified using predictors that are not constrained by temporalization.
+In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study.
+This is primarily because some evidence in Open Targets is not associated with publications.
+Ignoring that limitation is appropriate in this context.
 
@@
 This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ...
+(WIP)
 
@@
 Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}.
-It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance for the remaining \bottomRankingPct\%.
-We offer some evidence of this through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}).
+It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance more comprehensively.
+We offer some evidence of this through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}) and provides a particularly dramatic lift across the bottom \bottomRankingPct\% of TD pairs.
 While time is a necessary but not sufficient condition for clinical advancement of a TD pair, given that a single trial can take years to complete, it also correlates with how many distinct trials, drugs, companies, sponsors, investigators, etc.
 attempt to validate any one pair.
-We posit that this is crucial because even if a druggable, mechanistic link exists between a target and a disease, many (or even most) trials testing that link will fail for reasons unrelated to efficacy or safety of a drug \cite{Razuvayevskaya2023.02.07.23285407}.
+We posit that this is crucial because even if a druggable, mechanistic link exists between a target and a disease, many (or even most) trials testing that link will fail (or not even start) for reasons unrelated to efficacy or safety of a drug \cite{Razuvayevskaya2023.02.07.23285407}.
 Some of these reasons include 1) commercial factors like a lack of funding, pipeline reprioritization and competitive density, 2) administrative or logistical factors like a lack of enrollment, retention problems, poor trial design, drug supply chain shortages, epidemics (COVID-19) and 3) regulatory factors like legislative shifts or extensive regulatory approval delays.
@@
 It stands to reason then that predicting trial enrollment failures well in advance of when a trial is even conceived, i.e.
-for a target-disease pair, should be possible because recruitment in trials is determined, in part, by how prevalent a disease is, how debilitating it is, what patient demographics it inflicts, whether existing treatments exist for it and how successful recruitment for past trials (if any exist) has been -- all of which are measurable.
-Signals like this could be used to further improve performance in the bottom \bottomRankingPct\% of TD pairs since they better capture causal factors in non-biological trial failures that time under development -- the only non-biological feature in this study -- does not.
+for a target-disease pair, should be possible because recruitment in trials is determined, in part, by how prevalent a disease is, how debilitating it is, what patient demographics it inflicts, whether existing treatments exist for it and how successful recruitment for past trials has been -- all of which are measurable.
+Signals like this could be used to further improve performance since they better capture causal factors in non-biological advancement failures that time under development -- the only non-biological feature in this study -- does not.
 
@@
 It can also be argued that similarity between disease terms offers an important dimension for expanding evidence in a biologically meaningful way, regardless of technical mapping issues.
-This is taken to further extremes in studies like \cite{PMID:33262371} that propagates evidence using both measures of target similarity (from protein interaction networks) and measures of disease similarity (from ontologies and literature co-occurrence).
+This is taken to further extremes in studies like \cite{PMID:33262371} that propagates evidence using both measures of target similarity (e.g.
+using protein interaction networks) and measures of disease similarity (e.g.
+using ontologies and literature co-occurrence).
 We believe such expansion methods are very compatible with our approach by either using them to derive new predictors or to propagate predictions from models like these across networks or knowledge graphs.
 
-\begin{itemize}
-  \item Discuss the possibility to do prospective evaluation with OT snapshots
-  \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
-  \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
-  \item From \cite{PMID:33262371}: "It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices.
-Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid."
-\end{itemize}
+All results we have presented are based on quasi-prospective outcomes.
+As described in Section~\ref{sec:introduction}, this means that the outcomes are not truly observed in the future relative to the information used to predict them.
+Both outcomes and predictors are assigned timestamps corresponding to publication or database dates that approximately capture when the events associated with them actually occurred.
+The accuracy of that approximation is difficult to determine, and publication dates can be particularly inaccurate when they lag evidence discovery by months or years.
+This does not pose a risk for information leaks in a quasi-prospective formulation though, as long as the date associated with an event does not precede its occurrence.
+Nevertheless, it is possible that publication dates are simply entered into a database incorrectly, clinical trial registries inaccurately represent when trials began or ended, or that information from future events ends up associated with the wrong targets and/or diseases.
+These are only a few possible risks for information leaks, and though each is relatively rare on its own, their combined impact could be substantial.
+And as with all forecasting models, the accuracy of future predictions is reliant on the extent to which the underlying data generating process may either change or deviate from anticipated changes.
+Fortunately, the evidence we employ as predictors here comes predominantly from molecular interactions in humans, animals or cells that are likely to remain stable in time, or shift on evolutionary time scales that eclipse drug development timelines.
+This summary on the subject from \cite{PMID:33262371} is pertinent:
+
+\begin{quotation}
+  It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices.
+Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid.
+\end{quotation}
+
+Model performance evaluations based on prospective outcomes, as an alternative to quasi-prospective outcomes, offer a number of distinct advantages.
+For one, they mitigate nearly all the risks for information leaks mentioned previously by ensuring that all predictive information precedes clinical outcomes.
+They also offer a means to quantify those risks.
+An extension of this work might utilize the Open Targets database snapshots, dating back as far as 2016, to examine how much model performance differs between models trained on features from a snapshot and models trained on data from a current Open Targets release with features that are temporalized and then filtered to not exceed the year of the earlier snapshot.
+Similarly, that extension could examine how much model performance differs for a snapshot with and without longitudinal features by considering only outcomes that occurred after the snapshot.
+Such an approach could also be applied on a per-feature basis to gain more granular insight on which kinds of evidence pose the greatest risk.
+One clear disadvantage of prospective evaluation for this problem is that it requires years to accrue enough future clinical outcomes to be meaningful.
+This also means that predictors must omit that many years of accrued evidence as well, which can deflate performance compared to what is possible in the present day.
+Another disadvantage is that it is significantly more difficult to operationalize prospective evaluations, thereby implying that the decrease in risk of information leaks may be partially offset by an increase in technical risk.
 
@@
 
-\begin{itemize}
-  \item Discuss why all target prioritisation data fields are not used due to the potential leakage they may impose (e.g. target families, GO annotations, etc.)
-  \item Describe how therapeutic areas were selected based on having at least 100 TD pairs with target-disease-specific evidence of any kind and with explicit omissions: ("biological\_process", "pregnancy or perinatal disease", "injury, poisoning or other complication", "pregnancy or perinatal disease", "medical procedure", "infectious disease", "animal disease")
-  \item RDG models are fit using the \colorbox{Gainsboro}{Ridge} implementation from scikit-learn and GBM models are fit using the LightGBM \cite{LightGBM} algorithm
-  \item Cite supervenn
-  \item Training/evaluation results are limited by temporalization while the present-day predictions are not
-  \item OMIM is defined as EVA associations with publications
-  \item The "genetic\_association\_\_curated" field is a union of all genetic association sources other than "gene\_burden" and "ot\_genetics\_portal"
-  \item Mention that the scores for a TD pairs in a year are the maximum score for that source, not the harmonic sum
-  \item Mean imputation is used for target-specific features
-  \item Mention specifics on RDG and GBM implementations, i.e. lightGBM and scikit-learn
-  \item RDG is always trained on all features, but only applied to feature subsets where relevant
-\end{itemize}
+All data used in this study comes from Open Targets with one minor exception.
+Open Targets provides PubMed identifiers as provenance for evidence without supplying publication dates or years in all cases.
+These dates are taken from \cite{PMID:33763309} instead.
+Our primary results are based on the 23.12 Open Targets release, but we also use releases 23.09 and 23.06 as part of our sensitivity analysis in section \ref{sec:sensitivity}.
+
+The features we derive from Open Targets come primarily from a single target-disease dataset containing only "direct" evidence \cite{OTdirectVsIndirect}.
+These features are named according to the \colorbox{Gainsboro}{datasourceId} field within this dataset, and the values for those features are defined as the cumulative, maximum score observed when ordering those scores by the years associated with the evidence underlying them.
+We also add 2 genetic features that are not immediately available.
+The first, \colorbox{Gainsboro}{genetic\_association\_\_curated} is a union of all genetic association sources other than \colorbox{Gainsboro}{gene\_burden} and \colorbox{Gainsboro}{ot\_genetics\_portal}.
+The second, \colorbox{Gainsboro}{genetic\_association\_\_omim}, is defined as any EVA associations with a publication date.
+
+The target-specific features we derive come from the "target prioritisation" dataset \cite{OTtargetPrioritisation}.
+The only exception to this is a feature defining the maximum trial phase reached, which comes from the primary target-disease evidence dataset.
+Most scores in this dataset were not used because they cannot be temporalized.
+We make exceptions for those arising from large-scale, genome-wide assays like LOEUF from gnomAD \cite{PMID:32461654} and tissue expression scores from Human Protein Atlas \cite{PMID:25613900}.
+We retain a score derived from phenotype severity observed in mice knock-outs \cite{PMID:33231642} as well.
+All three of these feature sources are unique in that they are imputed differently from all other features when absent in model training.
+The "closed-world" assumption discussed in Section~\ref{sec:results_inflation} implies that a constant, zero imputation for most missing features is appropriate.
+This is not true for non-clinical, target-specific fields -- they are mean-imputed instead.
+
+All ridge regression models (e.g.
+RDG) are fit using the scikit-learn \colorbox{Gainsboro}{Ridge} implementation and all gradient-boosted machine models (e.g.
+GBM-T) are fit using the LightGBM \cite{LightGBM} algorithm.
+Ridge models are fit with very weak regularization (\colorbox{Gainsboro}{alpha=1e-6}) and LightGBM models are fit with default settings.
+Variants of both models are fit, in some cases, with non-negative monotonicity constraints.
+An important caveat unique to our use of the ridge regression models is that when report results for feature ablations, as in Section~\ref{sec:model_comparison}, all features are actually used to train the models before using only the unablated features to make predictions.
+
+Most results reported in this study consider only a subset of therapeutic areas within the Experimental Factor Ontology (EFO) \cite{PMID:20200009}.
+All therapeutic areas are retained if they possessed at least 100 TD pairs with target-disease-specific evidence of any kind within our evaluation dataset.
+We also explicitly omit the "biological\_process", "pregnancy or perinatal disease", "injury, poisoning or other complication", "pregnancy or perinatal disease", "medical procedure", "infectious disease", and "animal disease" therapeutic areas, as they are either not relevant or poorly represented.
+The "infectious disease" therapeutic area is a borderline case that we omit not because it is necessarily irrelevant, but because it has only 115 associated TD pairs.
+This makes it the most poorly represented therapeutic area that is possibly relevant.
+We remove it by explicit omission and use a 100 TD pair threshold rather than using a 115 TD pair threshold.
+
+\section{Limitations}
+
+There are a number of limitations and caveats with this study. These include:
+
+\begin{enumerate}
+  \item Open Targets does not currently provide drug approvals with associated dates.
+This is a part of the reason why we have chosen to focus only on advancement beyond phase 2, and not phase 3 as well.
+  \item Not all Open Targets evidence can be associated with publication dates.
+This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_opportunities}.
+  \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}.
+The score they provide is not longitudinal.
+We create an approximation of it over time based on the same weighting scheme across evidence types, but without a harmonic sum aggregation function.
+Maximum scores for the same target, disease, year and evidence type combination are used instead.
+  \item Estimating an absolute number of TD pairs associated with any criteria is difficult.
+In Section~\ref{sec:results_opportunities}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success.
+These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples.
+Defining a threshold for similarity to combat this problem is difficult and application-specific though, so we do not offer any solutions here.
+  \item The source of OMIM associations we use is derived from a subset of associations available in EVA.
+This subset is defined by associations with a linked publication.
+This is only an approximation, but our internal research shows that 98\% of EVA records with linked publications also have at least one OMIM submission.
+\end{enumerate}
+
+\pagebreak
 
@@
 
-\subsection{Features}
-
-\input{assets/feature_table.tex}
-
-\begin{figure}[!htb]
+\begin{figure}[H]
   \centering
@@
 
+\pagebreak
+
+\input{assets/feature_table.tex}
+
+\clearpage
+
 \section{Supplementary Material}
 
-\subsection{Performance}
+\pagebreak
 
@@
 
+\pagebreak
+
 \begin{figure}[H]
@@
 \end{figure}
+
+\pagebreak
 
@@
 
-\subsection{Predictions}
+\pagebreak
 
@@
 
-\subsection{Opportunities}
+\pagebreak
 
 \input{assets/tractability_bucket_table.tex}
+
+\clearpage
 
@@
 
+\pagebreak
+
 \subsection{Sensitivity}
+\label{sec:sensitivity}
 
@@
 \input{assets/sensitivity_configurations.tex}
+
+\clearpage
 
@@
 
+\pagebreak
+
 \bibliographystyle{unsrt}  

### 2024-04-11 paper/paper.tex
Add improved opportunity figure

@@
   \caption{
-    \textbf{Training and evaluation dataset summary statistics}.
+    \textbf{Training and evaluation dataset summary statistics}.\\\hspace{\textwidth}
+    \textbf{balance}: percentage of TD pairs that eventually advanced beyond phase 2 \\\hspace{\textwidth} 
+    \textbf{min\_year}: earliest year in which a TD pair has first advanced to phase 2 \\\hspace{\textwidth} 
+    \textbf{n\_targets}: number of unique targets \\\hspace{\textwidth} 
+    \textbf{n\_diseases}: number of unique diseases \\\hspace{\textwidth} 
+    \textbf{n\_pairs}: number of TD pairs (equal to number of records) \\\hspace{\textwidth} 
+    \textbf{n\_pairs\_wev}: number of TD pairs with directly associated evidence (i.e. not just target or disease evidence).
   }
@@
   \label{fig:relative_risk_dist_across_ta}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[H]
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
-  \caption{
-    \textbf{Performance by algorithm, constraint type and feature grouping}.  
-  }
-  \label{fig:relative_risk_model_features}
 \end{figure}

### 2024-04-11 analysis.py

@@ -2932,6 +2932,102 @@ def prepare_average_therapeutic_area_metrics(
     # fmt: on
 
 
+def prepare_opportunity_funnel(
+    opportunity_statistics: PandasDataFrame, color_fn: Callable[[str, float], str]
+) -> PandasDataFrame:
+    def _get_level_colors(levels: list[str]) -> dict[str, str]:
+        result = {}
+        for level in levels:
+            value = int(level.split(":")[1]) / 5.0
+            if "tractability" in level:
+                color = color_fn("Blues", value)
+            elif "development" in level:
+                color = color_fn("Reds", value)
+            else:
+                color = color_fn("Greens", value)
+            result[level] = color
+        return result
+
+    # fmt: off
+    df = (
+        opportunity_statistics
+        .pipe(lambda df: df[df["disease__therapeutic_area"] == "ALL"])
+        .pipe(lambda df: df[df["target__tractability__evidence"].isin(["ALL", "LOW", "MED", "HIGH"])])
+        .assign(benchmark=lambda df: df["benchmark_slug"])
+    )
+
+    facets = {
+        "stage": "current\ndevelopment\nstatus",
+        "threshold": "clinical\nadvancement\nprobability",
+        "tractability": "tractability by\nmodality",
+    }
+
+    stage_data = (
+        df
+        .pipe(lambda df: df[df["target__tractability__evidence"] == "ALL"])
+        .pipe(lambda df: df[df["target__tractability__modality"] == "ALL"])
+        .pipe(lambda df: df[df["target_disease__stage"] != "ALL"])
+        .pipe(lambda df: df[df["benchmark"] == "OTG"])
+        .pipe(assert_condition, lambda df: not df["target_disease__stage"].duplicated().any())
+        .assign(size=lambda df: df["model__n_pairs__over_threshold"].pipe(lambda s: s / s.sum()))
+        .assign(label=lambda df: df["target_disease__stage"])
+        .assign(level=lambda df: df["target_disease__stage"].map({
+            "NONE": 0, "Phase 1": 1, "Phase 2": 2, "Phase 3": 3, "Phase 4": 4,
+        }))
+        .assign(tick="stage")
+        .assign(facet=facets["stage"])
+    )
+    threshold_data = (
+        df
+        .pipe(lambda df: df[df["target__tractability__evidence"] == "ALL"])
+        .pipe(lambda df: df[df["target__tractability__modality"] == "ALL"])
+        .pipe(lambda df: df[df["target_disease__stage"] == "NONE"])
[…]

### 2024-04-12 paper/paper.tex
Finish opportunity funnel figure

@@
       \put(10.5,56){\line(1,0){.5}} % Top bar
-      \put(6,42.5){2x}
+      \put(6,42.5){\sffamily \fontsize{8pt}{12pt}\selectfont 2x}
       \put(30.5,22.25){\line(0,1){6.5}}
-      \put(23,25){1.5x}
+      \put(23,25){\sffamily \fontsize{8pt}{12pt}\selectfont 1.5x}
       % Ranking percentages
       \put(19,9){\line(0,1){4}}
-      \put(16,14){Top 1\%}
+      \put(15,14){\sffamily \fontsize{8pt}{12pt}\selectfont Top 1\%}
       \put(36.5,9){\line(0,1){4}}
-      \put(33.5,14){Top 2\%}
+      \put(32.5,14){\sffamily \fontsize{8pt}{12pt}\selectfont Top 2\%}
     \end{overpic}
@@
     \put(23,8){\line(0,1){2}}
-    \put(20,11){Top 1\%}
+    \put(20,11){\sffamily \fontsize{10pt}{12pt}\selectfont Top 1\%}
     \put(37,8){\line(0,1){2}}
-    \put(34,11){Top 2\%}
+    \put(34,11){\sffamily \fontsize{10pt}{12pt}\selectfont Top 2\%}
   \end{overpic}
@@
 The tractability buckets in Supplementary~Table~\ref{tab:tractability_buckets} provide \colorbox{Gainsboro}{HIGH}, \colorbox{Gainsboro}{MED}, and \colorbox{Gainsboro}{LOW} confidence ratings for each type of tractability evidence based on the priorities suggested in \cite{OTTractability}.
+
+\begin{figure}[!htb]
+  \centering
+  \captionsetup{width=.9\linewidth}
+  \begin{overpic}[width=1\textwidth]{opportunity_funnel.pdf}
+    \put(100, 53) {(a)}
+    \put(100, 41) {(b)}
+    \put(100, 19) {(c)}
+    \put(50, 51){\color{black}\vector(0, -1){5}}
+    \put(74, 39){\color{black}\vector(0, -1){4.5}}
+    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
+    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
+    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
+  \end{overpic}
+  \caption{
+    \textbf{Present-day target-disease pair opportunity funnel}.
+    ... 
+  }
+  \label{fig:opportunity_funnel}
+\end{figure}
 

### 2024-04-12 analysis.py

@@ -2025,6 +2025,14 @@ TRACTABILITY_BUCKETS = {
     },
 }
 
+TRACTABILITY_MODALITY_LABELS = {
+    # See https://github.com/chembl/tractability_pipeline_v2
+    "PR": "PROTAC",
+    "SM": "Small molecule",
+    "AB": "Antibody",
+    "OC": "Other",
+}
+
 
 def get_tractability_buckets(tractability: SparkDataFrame) -> PandasDataFrame:
     return (
@@ -2208,28 +2216,43 @@ def get_opportunity_statistics(
         .agg(
             F.count_distinct(F.col("pair")).alias("n_pairs"),
             F.array(
-                *[
-                    F.struct(
-                        F.lit(row.benchmark_name).alias("benchmark_name"),
-                        F.lit(row.benchmark_slug).alias("benchmark_slug"),
-                        F.count_distinct(
-                            F.when(
-                                F.col("prediction") >= row.model_prediction_threshold,
-                                F.col("pair"),
-                            )
-                        ).alias("model__n_pairs__over_threshold"),
-                        F.count_distinct(
-                            F.when(
-                                F.col(
-                                    f"target_disease__genetic_association__{row.benchmark_name}"
+                *(
+                    [
+                        F.struct(
+                            F.lit(row.benchmark_name).alias("benchmark_name"),
+                            F.lit(row.benchmark_slug).alias("benchmark_slug"),
+                            F.count_distinct(
+                                F.when(
+                                    F.col("prediction")
+                                    >= row.model_prediction_threshold,
+                                    F.col("pair"),
                                 )
-                                > 0,
-                                F.col("pair"),
-                            )
-                        ).alias("feature__n_pairs__exists"),
-                    )
-                    for row in (relative_risk_thresholds.itertuples())
-                ]
+                            ).alias("model__n_pairs__over_threshold"),
+                            F.count_distinct(
+                                F.when(
+                                    F.col(
+                                        f"target_disease__genetic_association__{row.benchmark_name}"
+                                    )
[…]

### 2024-04-12 paper/paper.tex
More figure captions

@@
 \usepackage{graphicx}       % graphics
+\usepackage{enumitem}       % lists
 \usepackage[table, x11names, svgnames]{xcolor}
@@
 \author{
-  Authors TBD \\
-  Related Sciences \\
-%   %% examples of more authors
-%    \And
-%   Author3 \\
-%   Affiliation \\
-%   Univ \\
-%   City\\
-%   \texttt{email@email} \\
+  Eric Czech\thanks{Correspondence to Eric Czech (eric@related.vc).} \hspace{0.2cm} 
+  Daniel Himmelstein \hspace{0.2cm} 
+  Rafal Wojdyla \hspace{0.2cm} 
+  \\ \\
+  \textbf{Jack Milwid} \hspace{0.2cm} 
+  \textbf{Adam Kolom} \hspace{0.2cm} 
+  \textbf{Jeff Hammerbacher}
+  \\ \\
+  Related Sciences LLC
 }
@@
 %% Body
-\setlength{\parindent}{20pt}
+\setlength{\parindent}{20pt} % Set paragraph indentation
+\setlength{\fboxsep}{1pt} % Set top and bottom padding around a \colorbox
 
@@
 
-It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:37803084,PMID:36963162}, and to a lesser extent, that the same may also be true for single-cell transcriptomic evidence \cite{Dann2024.04.04.24305313}.
+It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{PMID:24833294,PMID:35804044,Razuvayevskaya2023.02.07.23285407,PMID:37803084,PMID:36963162}, typically by a factor of at least 2x \cite{Minikel2023.06.23.23291765,Nelson2015-eg,King2019-rc,PMID:30652614,Sadler2023-xd}.
 This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}.
 It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}.
-To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates.
-This is not a focus of this work as we aim, instead, to establish a framework for identifying target-disease (TD) pairs with the very highest possible likelihood of success.
-We accomplish this by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in statistical models that are trained on past clinical outcomes, evaluated primarily on their performance at upper extremes of predictions and then deployed on undeveloped TD pairs to establish an evidence-based ranking pipeline.
-This method can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret.
-We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence, spanning hundreds or even thousands of putative orthogonal, causal predictors, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.
-This is the basis for our internal ranking platform, and such operational advantages have made it possible to reach or exceed the inclusivity of highly integrative methods without sacrificing interpretability, modularity and ease of deployment.
+In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines.
+This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.
+
+This method contrasts with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and often difficult to interpret.
+We believe a desirable middle ground between these two approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence from many sources to maximize inclusion of potentially causal, orthogonal predictive properties of targets and diseases, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.
 
 A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine.
-This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent otherwise.
-We use longitudinal evidence, i.e.
-evidence for which timing of its emergence can be determined, where appropriate when training and evaluating our methods before ultimately applying them to present-day evidence that is not temporalized.
-We discuss motivations, prior research and our own analysis on how important this problem is for each type of evidence in Section~\ref{sec:results_inflation}.
+This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent.
+By utilizing "temporalized" evidence -- i.e.
+evidence for which timing of its emergence can be determined -- and outcomes when training our methods before ultimately applying them to present-day evidence with no restrictions on timing, we reduce the risk of information leaks and inflation of performance estimates.
+We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.
 
@@
     \textbf{Present-day target-disease pair opportunity funnel}.
-    ... 
+    (a) All 2.53M TD pairs in Open Targets with evidence of some kind grouped by the maximum clinical trial phase reached for each as of the present day.
+The \colorbox{Gainsboro}{Undeveloped} pairs have never been tested in clinical trials.
+TD pair counts are shown in parentheses.
+Labels for phase 1, 3 and 4 trials are omitted for brevity.
+    (b) All 2.46M undeveloped TD pairs for which RDG model predictions exceed the thresholds determined by the corresponding genetic benchmarks in Figure~\ref{fig:relative_risk_by_limit}.
+The \colorbox{Gainsboro}{Medium} threshold was calibrated by genetic support across OTG (i.e.
+GWAS evidence).
+The associated RR of 2.07 implies that pairs with RDG predictions above this threshold are at least 2.07 more likely to advance clinically than those below the threshold.
+The same is true for the \colorbox{Gainsboro}{High} and \colorbox{Gainsboro}{Highest} thresholds, which correspond to thresholds determined by EVA and OMIM, respectively.
+These groupings are not mutually exclusive, e.g.
+the 20,821 TD pairs above the \colorbox{Gainsboro}{Medium} threshold are a superset of the pairs above the \colorbox{Gainsboro}{High} threshold.
+    (c) All 8,560 undeveloped TD pairs with an expected clinical advancement rate of at least 2.93x grouped by tractability modality and confidence level.
+The \colorbox{Gainsboro}{Other} modality corresponds to any modality not shown that is associated with a target tested in clinical trials previously.
+The \colorbox{Gainsboro}{High} confidence grouping for this category means those targets have approved drugs.
+For each other modality, the \colorbox{Gainsboro}{High} confidence level corresponds to \textbf{any} level of clinical validation for drugs of that modality, not just approval.
+See Supplementary~Table~\ref{tab:tractability_buckets} for the definition of all modality confidence levels.
   }
@@
 
-\begin{figure}[!htb]
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{opportunity_summary.png}
-  \caption{
-    \textbf{Present-day target-disease pair counts by stage, likelihood of advancement and tractability}.
-The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}.
-This corresponds to targets that have all been in clinical development already, except for the \textbf{OC} modality in which case it indicates that a target has been approved.
-  }
-  \label{fig:opportunity_summary}
-\end{figure}
+Figure~\ref{fig:opportunity_funnel} demonstrates this process visually, starting from the space of all present-day TD pairs.
+This funnel implies there are thousands of TD pairs that 1) have never been tested in clinical trials before, 2) have a 3x (RR=2.93) higher likelihood of advancing past phase 2 trials and 3) have a target that \textbf{has} been tested clinically (in at least phase 1 trials) for one or more of the
+modalities highlighted.
+In other words, there are $\sim$2,400 small-molecule-enabled, $\sim$1,400 antibody-enabled, and 14 PROTAC-enabled TD pairs at the bottom of the funnel where the implied opportunity is greatest.
+Examples of antibody-enabled TD pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.
 
-Figure~\ref{fig:opportunity_summary} shows the distribution of TD pair counts for select buckets across therapeutic areas as well as across the current maximum phase reached for any one pair.
-We find that there are $\sim$2,400 small-molecule-enabled, $\sim$1,400 antibody-enabled, and 14 PROTAC-enabled TD pairs with a probability of advancement that is nearly 3x other TD pairs based on the EVA threshold RR=2.93 in Figure~\ref{fig:relative_risk_by_limit}.
-Top antibody-enabled pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.
+For similar statistics by therapeutic area, rather than across all of them as in Figure~\ref{fig:opportunity_funnel}, see Supplementary~Figure~\ref{fig:opportunity_summary}.
+This also shows TD pair counts for trial phases omitted from Figure~\ref{fig:opportunity_funnel}.
 
-We also note that these TD pairs are identified using predictors that are not constrained by temporalization.
+Lastly, we note that these TD pairs are identified using predictors that are not constrained by temporalization.
 In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study.
 This is primarily because some evidence in Open Targets is not associated with publications.
-Ignoring that limitation is appropriate in this context.
+Avoiding that limitation is appropriate in this context.
 
@@
 
-\begin{enumerate}
+\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
 \item Mendelian randomization research is biased towards cardiovascular diseases as they have a disproportionate number of known, modifiable exposures \cite{PMID:36736292}
@@
 
-We also note that the skew in basic drug target research towards those that already have rich annotations and well characterized molecular function \cite{PMID:29358745} as well as the disproportionate representation of particular target families in pharma R\&D pipelines \cite{PMID:27910877,PPR:PPR7029} and the fact that literature is well known to be biased away from negative results in general \cite{PMID:32893970} are all problematic.
+We also note that the skew in basic drug target research towards those that already have rich annotations and well characterized molecular function \cite{PMID:29358745}, the disproportionate representation of particular target families in pharma R\&D pipelines \cite{PMID:27910877,PPR:PPR7029} and the fact that literature is well known to be biased away from negative results in general \cite{PMID:32893970} are all problematic.
 
@@
 
-\begin{itemize}
+\begin{itemize}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
 \item \(A\) is the event that evidence for a TD pair arises after its first early-stage (phase 1 or 2) trial rather than before
@@
   \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{evidence_inflation.pdf}
+  \begin{overpic}[width=1\textwidth]{evidence_inflation.pdf}
+    \put(73, 42) {(a)}
+    \put(73, 16) {(b)}
+  \end{overpic}
   \caption{
-    \textbf{Clinical success drives evidence discovery}.  
+    \textbf{Inflation risk by feature}.  
+    (a) Feature "inflation risk", for more frequent features, reported as relative risk point estimates (dots) and Katz 99\% confidence intervals (bars).
+These features are "frequent" in the sense that they have enough TD pairs where evidence emerged before \textbf{and} after a phase 2 transition to power a relatively narrow confidence interval.
+    (b) Rare feature inflation risk with the same definition as (a) and a wider range on the x-axis.
+All point estimates and confidence intervals are clamped to the range (0, 10) to minimize scale distortion from extreme estimates.
   }
@@
 Pushing that even further to 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively.
-These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence, however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e.
+These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e.
 untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.
 
-% TODO: complete this section?
-Another way to quantify the impact that employing superior target prioritization strategies could have is to consider what would happen if this was done across the entire pharmaceutical development industry.
+Another way to quantify the impact that employing superior target prioritization strategies could have, is to consider what would happen if this was done across the entire pharmaceutical development industry.
 Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year.
 This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ...
-(WIP)
+(TODO: finish with Adam)
 
@@
 
-\begin{quotation}
+\begin{quote}
   It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices.
 Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid.
-\end{quotation}
+\end{quote}
 
@@
 
-The target-specific features we derive come from the "target prioritisation" dataset \cite{OTtargetPrioritisation}.
+The target-specific features we derive come from the "target prioritisation" dataset \cite{OTtargetPrioritisation} in Open Targets.
 The only exception to this is a feature defining the maximum trial phase reached, which comes from the primary target-disease evidence dataset.
@@
 
-\begin{enumerate}
+\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
   \item Open Targets does not currently provide drug approvals with associated dates.
@@
     \textbf{Training and evaluation dataset summary statistics}.\\\hspace{\textwidth}
-    \textbf{balance}: percentage of TD pairs that eventually advanced beyond phase 2 \\\hspace{\textwidth} 
-    \textbf{min\_year}: earliest year in which a TD pair has first advanced to phase 2 \\\hspace{\textwidth} 
-    \textbf{n\_targets}: number of unique targets \\\hspace{\textwidth} 
-    \textbf{n\_diseases}: number of unique diseases \\\hspace{\textwidth} 
-    \textbf{n\_pairs}: number of TD pairs (equal to number of records) \\\hspace{\textwidth} 
-    \textbf{n\_pairs\_wev}: number of TD pairs with directly associated evidence (i.e. not just target or disease evidence).
+    \colorbox{Gainsboro}{balance}: percentage of TD pairs that eventually advanced beyond phase 2 \\\hspace{\textwidth} 
+    \colorbox{Gainsboro}{min\_year}: earliest year in which a TD pair has first advanced to phase 2 \\\hspace{\textwidth} 
+    \colorbox{Gainsboro}{n\_targets}: number of unique targets \\\hspace{\textwidth} 
+    \colorbox{Gainsboro}{n\_diseases}: number of unique diseases \\\hspace{\textwidth} 
+    \colorbox{Gainsboro}{n\_pairs}: number of TD pairs (equal to number of records) \\\hspace{\textwidth} 
+    \colorbox{Gainsboro}{n\_pairs\_wev}: number of TD pairs with directly associated evidence (i.e. not just target or disease evidence).
   }
@@
 \input{assets/feature_table.tex}
+
+\clearpage
+
+\input{assets/tractability_bucket_table.tex}
 
@@
     \textbf{Performance of individual features and predictive scores as measured by relative risk}.
+    RR estimates are shown as dots with bars indicating 90\% Katz confidence intervals.
     RDG model results denoted by \colorbox{Gainsboro}{rdg@N} indicate performance for the \colorbox{Gainsboro}{N} top TD pairs.
@@
   \caption{
-    \textbf{Relative risk scores for target/disease features}.
+    \textbf{Relative risk scores for target-only and disease-only features}.
+    RR estimates are shown as dots with bars indicating 90\% Katz confidence intervals.
     The features ending with \colorbox{Gainsboro}{qtl\_Q} denote binary indicators constructed from cases where the feature meets or exceeds quantile \colorbox{Gainsboro}{Q} of its distribution.
 The features ending with \colorbox{Gainsboro}{gte\_X} denote indicators for when the feature meets or exceeds a specific value \colorbox{Gainsboro}{X}.
+The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
   }
@@
     \textbf{Relative risk scores by method, benchmark and therapeutic area}.
-    The \textbf{average} therapeutic area indicates mean values across all others except for \textbf{all}, which is an ungrouped estimate across all diseases regardless of therapeutic area.
+    The \colorbox{Gainsboro}{average} therapeutic area denotes mean values across all other therapeutic areas except for \colorbox{Gainsboro}{all}, which is an ungrouped estimate across all diseases regardless of therapeutic area.
+The \colorbox{Gainsboro}{n\_pairs} field indicates the number of TD pairs used to compute the RR estimate in each cell.
+The \colorbox{Gainsboro}{OTS@N} and \colorbox{Gainsboro}{RDG@N} columns indicate RR estimates for the top \colorbox{Gainsboro}{N} TD pairs from the OTS and RDG methods, respectively.
   }
@@
     \textbf{Relative risk distributions across select therapeutic areas}.
-    P-values are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR averages across therapeutic areas exceed OTS averages.
+    Relative risk point estimates for each of 13 therapeutic areas are shown as dots with boxplots underneath summarizing their distribution.
+P-values on the bottom margin of the top facet are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR average across therapeutic areas exceeds the OTS average at a given number of top ranking TD pairs (on the x-axis).
   }
@@
     Feature contributions are shown as the product of their underlying values and the RDG coefficients.
-The \textbf{advanced} field indicates whether the associated TD pair advanced beyond phase 2 as of 2024.
+The \colorbox{Gainsboro}{advanced} field indicates whether the associated TD pair advanced beyond phase 2 as of 2024.
+The \colorbox{Gainsboro}{prediction} field shows the final RDG model prediction for each case.
   }
@@
 \end{figure}
-
-\pagebreak
-
-\input{assets/tractability_bucket_table.tex}
 
@@
   \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{opportunity_summary.png}
+  \caption{
+    \textbf{Present-day target-disease pair counts by stage, likelihood of advancement and tractability}.
+The \colorbox{Gainsboro}{stage} panel contains counts by maximum trial phase reached, the \colorbox{Gainsboro}{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \colorbox{Gainsboro}{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a \colorbox{Gainsboro}{HIGH} tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}.
+See Figure~\ref{fig:opportunity_funnel} for more details and a more detailed visual summary of these same results for the \colorbox{Gainsboro}{ALL} therapeutic area, which simply indicates counts across all diseases regardless of therapeutic area.
+  }
+  \label{fig:opportunity_summary}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
   \includegraphics[width=.9\textwidth]{top_opportunity_predictions.png}
   \caption{
-    \textbf{Top ranked undeveloped, tractable target-disease pairs}.
-The highest scoring TD pairs per the RDG model that have not entered clinical trials despite having a target that has been in trials of an antibody-based drug.
-All values shown other than \textbf{prediction} are raw feature values, unweighted by RDG coefficients.
+    \textbf{Top ranked undeveloped, antibody-enabled target-disease pairs}. 
+    Present-day, undeveloped, antibody-enabled TD pairs with the highest RDG model predictions (shown in the \colorbox{Gainsboro}{prediction} column) are presented here.
+These pairs are "undeveloped" in the sense that they have no known clinical trials.
+Furthermore, they are "antibody-enabled" in the sense that the targets for each pair have been previously subjected to antibody-based drugs in clinical trials.
+More details and context are provided in Figure~\ref{fig:opportunity_funnel}.
   }

### 2024-04-12 analysis.py

@@ -27,6 +27,7 @@ from pyspark.sql import functions as F
 from pyspark_gcs import get_gcs_enabled_config
 from scipy.stats import wilcoxon
 from scipy.stats.contingency import relative_risk
+from scipy.stats import fisher_exact
 from sklearn.base import BaseEstimator
 from sklearn.compose import ColumnTransformer
 from sklearn.discriminant_analysis import StandardScaler
@@ -1413,6 +1414,91 @@ def get_relative_risk_by_therapuetic_area(
     return pd.concat(dfs, axis=0, ignore_index=True).pipe(
         assert_condition,
         lambda df: not df[["therapeutic_area_name", "method"]].duplicated().any(),
+    )
+
+
+###############################################################################
+# Inflation
+###############################################################################
+
+
+def get_feature_inflation_statistics(
+    advancement_statistics: PandasDataFrame,
+) -> PandasDataFrame:
+    def _get_relative_risk(
+        r: pd.Series, confidence_level: float = 0.99
+    ) -> dict[str, float]:
+        counts = dict(
+            exposed_pos=int(r[("after", "advanced")]),
+            exposed_neg=int(r[("before", "advanced")]),
+            control_pos=int(r[("after", "stalled")]),
+            control_neg=int(r[("before", "stalled")]),
+        )
+        n_exposed = counts["exposed_pos"] + counts["exposed_neg"]
+        n_control = counts["control_pos"] + counts["control_neg"]
+        if n_exposed == 0 or n_control == 0:
+            return dict()
+        return dict(
+            relative_risk_low=(
+                rr := relative_risk(
+                    counts["exposed_pos"],
+                    n_exposed,
+                    counts["control_pos"],
+                    n_control,
+                )
+            )
+            .confidence_interval(confidence_level=confidence_level)
+            .low,
+            relative_risk_high=rr.confidence_interval(
+                confidence_level=confidence_level
+            ).high,
+            relative_risk=rr.relative_risk,
+        )
+
+    return (
+        advancement_statistics.pipe(lambda df: df[df["progress"] != "none"])
+        .set_index(["feature_name", "emerged", "progress"])["n_pairs"]
+        .unstack(["emerged", "progress"])
+        .fillna(0)
+        .astype(int)
+        .pipe(
+            lambda df: (
+                pd.concat(
+                    [
+                        df,
+                        pd.DataFrame(
+                            {
+                                (c, "fraction"): df[c]["advanced"] / df[c].sum(axis=1)
[…]

### 2024-04-13 paper/paper.tex
Merge in more edits

@@
 \usepackage[section]{placeins}
+\usepackage[symbol]{footmisc}
 \usepackage{setspace}
@@
   Eric Czech\thanks{Correspondence to Eric Czech (eric@related.vc).} \hspace{0.2cm} 
+  Rafal Wojdyla \hspace{0.2cm} 
   Daniel Himmelstein \hspace{0.2cm} 
-  Rafal Wojdyla \hspace{0.2cm} 
   \\ \\
@@
 This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases.
-We demonstrate a novel statistical learning framework utilizing both constrained and unconstrained linear and tree models, that can predict clinical advancement for the top \topRankingPct\% of target-disease pairs 1.5-2x more effectively than Open Targets composite scores as well as common measures of genetic support that have already been established -- and observed in this study -- to confer a 2x higher likelihood of success (a 3-4x cumulative improvement).
-Utilizing a subset of our biomedical evidence base, non-negative linear models can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify target-disease pairs with high probabilities of advancing beyond phase 2 trials, and uncover a range of previously undeveloped target-disease pairs poised for clinical success.
+We demonstrate a novel statistical learning framework for identifying the top \topRankingPct\% of target-disease pairs that are 4-6x more likely to advance beyond phase 2 trials.
+This framework is 1.5-2x more effective than an Open Targets composite score based on the same set of evidence.
+It is also 2x more effective than a common measure for genetic support that has been observed previously, as well as in this study, to confer a 2x higher likelihood of success.
+Utilizing a subset of our biomedical evidence base, non-negative linear models resulting from this framework can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify previously undeveloped target-disease pairs poised for clinical success.
 In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different type of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success.
@@
 
-While the use of longitudinal evidence is very rare in studies on target prioritization algorithms, it is not at nearly as rare in studies that attempt to predict clinical trial outcomes using a variety of biological evidence sources, clinical statistics and other organizational or behavioral factors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}.
-The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets or diseases, trial sponsor track records, eventual patient enrollment, etc.
-constitute clear information leaks otherwise.
-This is discussed in some depth in \cite{PMID:37483175} which notes several studies that do not account for this problem before drawing a clear distinction between quasi-prospective and prospective problem formulations.
-The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until sufficient time has elapsed for more outcomes to occur.
-Nomenclature for these formulations is conflicting though, where this definition of a quasi-prospective formulation is deemed entirely prospective in some cases, e.g.
-\cite{PMID:37225853}.
+Solutions to such problems are common, but not fully addressed even in precedent studies by other groups that attempt to predict clinical trial outcomes using similarly temporalized predictors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}.
+The need for longitudinal features is most clear in settings where the inclusion of predictors like historical success rates for targets or diseases, trial sponsor track records, eventual patient enrollment, or other time-based events and outcomes, would otherwise constitute information leaks.
+This is discussed in some depth in \cite{PMID:37483175} which notes several studies that do not account for this problem, and presents both quasi-prospective as well as prospective results.
+The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until years later.
+Nomenclature for these formulations is conflicting though, where this definition of a quasi-prospective design is deemed entirely prospective in some cases (e.g.
+\cite{PMID:37225853}).
 To avoid any potential confusion, our formulation in this study is quasi-prospective.
-The distinct advantages and disadvantages of each formulation provide a complementary risk/reward trade-off that we capitalize on in our internal platform, and we assert that the distinction between them is important.
-Interestingly, the existence of Open Targets snapshots dating back to at least 2016 would enable the extension of this work to a prospective formulation as well.
+The distinct advantages and disadvantages of each design provide a complementary risk/reward trade-off, and we assert that the distinction between them is important.
 This is discussed more in Section~\ref{sec:discussion}.
 
-The preceding works discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success and 2) clinical trial outcome prediction models.
-Both are measured against the same outcomes and an important distinction between them lies in how the prioritization methods are {\bf not} directly optimized for those outcomes while the outcome prediction methods are.
-In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that could be present well in advance of any drug program or individual trial.
+Precedent studies and methods discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success, and 2) clinical trial outcome prediction models.
+Both are measured against the same outcomes and an important distinction between them lies in how prioritization methods are \textbf{not} directly optimized for those outcomes while outcome prediction methods are.
+In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial.
 We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS.
-Finally, we examine how many present-day target-disease pairs are undeveloped (i.e.
-have never been in clinical trials), have a tractable target and are likely to see success rates matching or exceeding those benchmarks.
+Finally, we examine how many present-day target-disease pairs are likely to see success rates matching or exceeding those calibrated benchmarks, that are also undeveloped (i.e.
+have never been in clinical trials) and have a tractable target.
 
@@
 
-In order to model clinical advancement for target-disease pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with any one TD pair as indicated by the presence of a later-stage trial.
-All results to follow consider only advancement beyond phase 2 due to limitations described in Section~\ref{sec:discussion}.
-This binary outcome is then predicted based on a list of features shown in Table~\ref{tab:features}.
-Information for each of these features is only used when it was published before the year \textbf{prior} to the first phase 2 trial observed, with an exception for genetic evidence discussed in Section~\ref{sec:results_inflation}.
-A training dataset is then formed by including only TD pairs where this first phase 2 year is between 1990 and 2015.
-The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete.
+In order to model the probability of clinical success, or "clinical advancement" for target-disease (TD) pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with a single TD pair.
+We consider an advancement to have occurred when a later-stage trial is observed for the same pair.
+When no later-stage trial is observed, no advancement is assumed and this binary condition provides the outcome to predicted using the features shown in Table~\ref{tab:features}.
+Due to the lack of timestamped drug approvals in Open Targets, all study results presented here consider only advancement beyond phase 2.
+This limitation is discussed further in Section~\ref{sec:limitations}.
+
+Information for each of the features in Table~\ref{tab:features} that are denoted as \colorbox{Gainsboro}{temporal} is only used when it was published in a year \textbf{prior} to the first phase 2 trial for a TD pair.
+Genetic evidence is a notable exception to this rule, and motivations for that exception are detailed in Section~\ref{sec:results_inflation}.
+A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015.
+The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \underline{2 year} offset from the present year (2024) to allow enough time for some trials to complete.
 While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}.
-This means a substantial fraction of outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation.
+This means a substantial fraction of recent outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation.
 The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.
@@
 
-The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor.
+The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor -- a subset of our full productionized feature set.
 These are listed in Table~\ref{tab:features}.
@@
 This choice was motivated in part by empirical results presented in Section~\ref{sec:results_inflation} and assumptions described more in Section~\ref{sec:methods}.
-Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OTtargetPrioritisation} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (also described in Section~\ref{sec:methods}).
+Simply put, the data in this study combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OTtargetPrioritisation} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (also described in Section~\ref{sec:methods}).
 
@@
 
-We train a variety of models including constrained and unconstrained linear and tree models.
-The constrained variants of these models force effects of all features to increase monotonically, i.e.
-all effects are constrained to be non-negative.
+We develop our predictive framework using two primary types of models: linear models, where relationships between features and outcomes are assumed to be linear, and gradient-boosted tree models, which can capture non-linear relationships.
+Both models were tested in constrained formats, ensuring all feature effects are non-negative, and unconstrained formats, allowing for negative relationships.
 This is possible with no underlying feature transformations because all scores in Open Targets are constructed such that higher scores are presumed to be advantageous.
+We also note that neural models are a significant omission from this framework.
+While deep learning methods for tabular data are competitive with gradient-boosted tree models \cite{mcelfresh2023neural, gorishniy2023revisiting, somepalli2021saint}, they are rarely superior without transfer learning \cite{mcelfresh2023neural, levin2023transfer, hollmann2023tabpfn}.
+They also have a number of operational drawbacks like necessitating GPUs and having less standard hyperparameter tuning procedures.
+Additionally, the data in this proof-of-concept is limited to Open Targets evidence scores that already reflect a large degree of feature engineering, thereby limiting the extent to which deep learning networks can identify more granular predictive signals.
 
-We also apply these models to our evaluation dataset using several feature ablations in order to assess the value of groups of related features.
+In order to assess the value of groups of related features, we also apply these models to our evaluation dataset using several feature ablations.
 We refer to a "core" feature set consisting of all features listed in Table~\ref{tab:features} except for the sole feature capturing the time since a target-disease pair first entered phase 2 trials (i.e.
@@
 \begin{itemize}
-  \item \textbf{RDG}: Constrained, L2-regularized linear regressor (a.k.a. "Ridge regressor") fit with all core features
-  \item \textbf{RDG-T}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
-  \item \textbf{RDG-X}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
-  \item \textbf{GBM}: Constrained, gradient-boosted machine with fit with all core features
-  \item \textbf{GBM-T}: \textbf{GBM} fit with all features instead of only core features
+  \item \textbf{RDG}\footnote[2]{\label{ours}These models are constructed in this study rather than being defined by an external source.}: Constrained, L2-regularized linear regressor (a.k.a.
+"Ridge regressor") fit with all core features
+  \item \textbf{RDG-T}\textsuperscript{\ref{ours}}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
+  \item \textbf{RDG-X}\textsuperscript{\ref{ours}}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
+  \item \textbf{GBM}\textsuperscript{\ref{ours}}: Constrained, gradient-boosted machine with fit with all core features
+  \item \textbf{GBM-T}\textsuperscript{\ref{ours}}: \textbf{GBM} fit with all features instead of only core features
   \item \textbf{OTS}: Open Targets composite score
@@
 In order to compare these models to an Open Targets composite score (OTS), we use an equally weighted sum of all scores from individual sources except for those assigned lower weights in \cite{OTweights}.
-Scores from these sources are multiplied by the corresponding weight before being summed and only the TD-specific features of Table~\ref{tab:features} are used.
-Neither the target/disease specific features nor the time since phase 2 transition feature are included in this calculation.
+Scores from these sources are multiplied by the corresponding weight before being summed and only the target-disease-specific features of Table~\ref{tab:features} are used.
+Neither the target-specific and disease-specific features nor the time since phase 2 transition feature are included in this calculation.
 
@@
     \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
-    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
+    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x+ more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
   \end{overpic}
@@
 \section{Limitations}
+\label{sec:limitations}
 

### 2024-04-13 analysis.py

@@ -2537,6 +2537,29 @@ def set_default_table_styles(styler: Styler, height_px: int = 250) -> Styler:
     )
 
 
+# TODO: delete if custom colorscales don't end up being used
+# def get_base_colors() -> dict[str, tuple[float, float, float]]:
+#     return {
+#         "red": (252, 37, 92),
+#         "cyan": (28, 195, 227),
+#         "purple": (97, 84, 255),
+#         "yellow": (226, 204, 22),
+#         "grey": (127, 143, 186),
+#     }
+
+# def get_base_hex_colors() -> dict[str, str]:
+#     return {name: mcolors.to_hex(tuple(c/255 for c in rgb)) for name, rgb in get_base_colors().items()}
+
+# def get_colormaps() -> dict[str, mcolors.LinearSegmentedColormap]:
+#     cmaps = {}
+#     for name, rgb in get_base_colors().items():
+#         rgb = np.array(rgb)
+#         colors = [rgb + (255 - rgb) * i / 10 for i in range(11)]
+#         gradient = [tuple(c/255 for c in color) for color in colors[::-1]]
+#         cmaps[name] = mcolors.LinearSegmentedColormap.from_list(f"_{name.title()}s", gradient, N=256)
+#     return cmaps
+
+
 def get_assets_directory() -> Path:
     path = Path(__file__).parent.resolve() / "paper" / "assets"
     if not path.exists():

### 2024-04-13 paper/paper.tex
Second draft complete

@@
 \usepackage{enumitem}       % lists
+\usepackage{soul}           % underlines
 \usepackage[table, x11names, svgnames]{xcolor}
@@
 \usepackage{overpic}
+\usepackage{titlesec}
 
@@
 \setlength{\fboxsep}{1pt} % Set top and bottom padding around a \colorbox
+\titlespacing\section{0pt}{8pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt} % Prevent text after section headings from floating to bottom of page
 
@@
 This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases.
-We demonstrate a novel statistical learning framework for identifying the top \topRankingPct\% of target-disease pairs that are 4-6x more likely to advance beyond phase 2 trials.
+We demonstrate a novel statistical learning framework for identifying the top \topRankingPct\% of target-disease pairs that are as much as 4-5x more likely to advance beyond phase 2 trials.
 This framework is 1.5-2x more effective than an Open Targets composite score based on the same set of evidence.
@@
 Utilizing a subset of our biomedical evidence base, non-negative linear models resulting from this framework can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify previously undeveloped target-disease pairs poised for clinical success.
-In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different type of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success.
+In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different types of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success.
 To ease reproduction and deployment, no data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
@@
 It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}.
+\ul{To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates}.
 In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines.
@@
 
-A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine.
+\ul{A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine}.
 This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent.
@@
 Both are measured against the same outcomes and an important distinction between them lies in how prioritization methods are \textbf{not} directly optimized for those outcomes while outcome prediction methods are.
-In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial.
+\ul{In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial}.
 We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS.
@@
 A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015.
-The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \underline{2 year} offset from the present year (2024) to allow enough time for some trials to complete.
+The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \ul{2 year} offset from the present year (2024) to allow enough time for some trials to complete.
 While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}.
@@
 \label{sec:features}
-
-The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor -- a subset of our full productionized feature set.
-These are listed in Table~\ref{tab:features}.
-The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods.
-This choice was motivated in part by empirical results presented in Section~\ref{sec:results_inflation} and assumptions described more in Section~\ref{sec:methods}.
-Simply put, the data in this study combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OTtargetPrioritisation} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (also described in Section~\ref{sec:methods}).
 
@@
 
+The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor -- a subset of our full productionized feature set.
+These are listed in Table~\ref{tab:features}.
+The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods.
+This choice was motivated in part by empirical results presented in Section~\ref{sec:results_inflation} and assumptions described more in Section~\ref{sec:methods}.
+Simply put, the data in this study combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OTtargetPrioritisation} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (also described in Section~\ref{sec:methods}).
+
 The combination of target, disease and target-disease features creates sparsity patterns that are important to understand before interpreting models built from it.
-Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=\evaluationDatasetPairCount), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
+Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=\evaluationDatasetPairCount), slightly less than 2\% of them (1.89\%) ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs.
 These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases.
@@
 Taken together, the paucity of TD pair evidence and the abundance of prior target or disease clinical validation should be considered carefully when interpreting predictive performance.
-TD pairs predicted to be highly likely to advance clinically despite a lack of target-disease-specific evidence are entirely plausible, but the value of such a prediction is dependent upon the application.
+TD pairs predicted to be highly likely to advance clinically despite a lack of target-disease-specific evidence are entirely plausible, but the utility of such a prediction may vary depending upon the application.
 We choose to minimize the influence of these cases in our study by focusing on rankings within therapeutic areas that do not extend beyond the number of TD pairs with direct evidence of some kind, as discussed in Section~\ref{sec:methods}.
@@
 \begin{itemize}
-  \item \textbf{RDG}\footnote[2]{\label{ours}These models are constructed in this study rather than being defined by an external source.}: Constrained, L2-regularized linear regressor (a.k.a.
-"Ridge regressor") fit with all core features
-  \item \textbf{RDG-T}\textsuperscript{\ref{ours}}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
-  \item \textbf{RDG-X}\textsuperscript{\ref{ours}}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
-  \item \textbf{GBM}\textsuperscript{\ref{ours}}: Constrained, gradient-boosted machine with fit with all core features
-  \item \textbf{GBM-T}\textsuperscript{\ref{ours}}: \textbf{GBM} fit with all features instead of only core features
-  \item \textbf{OTS}: Open Targets composite score
+  \item \textbf{RDG}: Constrained, L2-regularized linear regressor (a.k.a. "Ridge regressor") fit with all core features
+  \item \textbf{RDG-T}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
+  \item \textbf{RDG-X}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
+  \item \textbf{GBM}: Constrained, gradient-boosted machine with fit with all core features
+  \item \textbf{GBM-T}: \textbf{GBM} fit with all features instead of only core features
 \end{itemize}
@@
 We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement.
-This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult.
 See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model.
@@
 
+\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
+  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
+  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
+  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
+  \item Linear models outperform gradient-boosting models by nearly all measures
+  \item Constrained linear models outperform unconstrained linear models by nearly all measures
+\end{enumerate}
+
 \begin{figure}[!htb]
@@
 
-\begin{enumerate}
-  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
-  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
-  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
-  \item Linear models outperform gradient-boosting models by nearly all measures
-  \item Constrained linear models outperform unconstrained linear models by nearly all measures
-\end{enumerate}
-
 We conclude from these results that constrained linear models are an optimal choice for this problem due both to their greater performance and interpretability.
@@
 
-\subsection{Effects}
+\subsection{Feature effects}
 \label{sec:effects}
@@
 EVA, Orphanet, UniProt, Genomics England, ClinGen and gene2phenotype, via the \colorbox{Gainsboro}{curated} \vspace*{0mm} feature.
-Notably, literature and target/disease specific clinical features also have substantial effects, followed by indicators of animal evidence and target genetic constraint / expression specificity.
+Notably, literature and target/disease specific clinical features also have substantial effects, followed by indicators of animal evidence and target genetic constraint or expression specificity.
 Any features not shown were deflated to have no effect, which is possible in this model due to the non-negativity constraint.
@@
 \label{sec:results_opportunities}
+
+Utilizing our predictive models, we systematically identify target-disease pairs that have not yet been explored in clinical trials but exhibit a high likelihood of successful outcomes.
+This process involved ranking pairs based on their predicted success rates and filtering those with substantial supporting evidence but no prior clinical development attempts.
 
@@
 
+To address potential biases and inflation in our dataset, our methodology specifically considers the temporal emergence of evidence.
+By doing so, we ensure our predictions are grounded in data reflective of genuine biomedical insights, rather than artefacts of research focus or historical data accumulation trends.
+
 Like most studies of this kind, we assume a "closed-world" \cite{Paliwal2020-hr} over the space of target-disease pairs and any evidence between them.
@@
 While it is not possible to address all of these issues, we emphasize that there is a clear pattern across the examples in the list above in that they require \textbf{past} clinical successes and/or failures to arise in the first place.
-This suggests that accounting for when evidence first emerged would limit the extent of these problems.
+\ul{This suggests that accounting for when evidence first emerged would limit the extent of these problems}.
 We do so in this study based solely on publication dates associated with any one piece of information linking target-disease pairs.
-This also offers a novel opportunity to attempt to quantify what kind of evidence suffers most from these biases.
+\ul{This also offers a novel opportunity to attempt to quantify what kind of evidence suffers most from these biases}.
 Figure~\ref{fig:evidence_inflation} presents results for this based on a relative risk statistic defined as:
@@
 
-We have demonstrated that simple machine learning methods applied to longitudinal biomedical evidence from many sources can be used to predict clinical outcomes for combinations of drug targets and diseases, without knowledge of molecular properties or trial design details.
+We have demonstrated that simple machine learning methods applied to longitudinal biomedical evidence from many sources can be used to identify combinations of drug targets and diseases that are 4-5x more likely to advance past phase 2 trials, and that this can be done without knowledge of molecular properties or trial design details.
 We have also shown that these methods are more precise in the extremes of their predictions than composite, heuristic scores like those from Open Targets.
@@
 If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug.
-Pushing that even further to 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively.
+Pushing that even further to a 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively.
 These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e.

### 2024-04-14 paper/paper.tex
Minor updates

@@
 Pushing that even further to a 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively.
-These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e.
-untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.
+These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} that there are more than 8,000 such opportunities (i.e.
+untested, present-day TD pairs) where at least a 3x improvement in phase 2 success rates could be expected.
 
@@
   \captionsetup{width=.9\linewidth}
+  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
+  \caption{
+    \textbf{Relative risk distributions across select therapeutic areas}.
+    Relative risk point estimates for each of 13 therapeutic areas are shown as dots with boxplots underneath summarizing their distribution.
+P-values on the bottom margin of the top facet are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR average across therapeutic areas exceeds the OTS average at a given number of top ranking TD pairs (on the x-axis).
+  }
+  \label{fig:relative_risk_dist_across_ta}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[H]
+  \centering
+  \captionsetup{width=.9\linewidth}
   \includegraphics[width=1\textwidth]{relative_risk_by_ta.png}
@@
   \label{fig:relative_risk_by_ta}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[H]
-  \centering
-  \captionsetup{width=.9\linewidth}
-  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
-  \caption{
-    \textbf{Relative risk distributions across select therapeutic areas}.
-    Relative risk point estimates for each of 13 therapeutic areas are shown as dots with boxplots underneath summarizing their distribution.
-P-values on the bottom margin of the top facet are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR average across therapeutic areas exceeds the OTS average at a given number of top ranking TD pairs (on the x-axis).
-  }
-  \label{fig:relative_risk_dist_across_ta}
 \end{figure}

### 2024-04-16 paper/paper.tex
Minor updates

@@
 A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015.
-The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \ul{2 year} offset from the present year (2024) to allow enough time for some trials to complete.
+The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete.
 While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}.
@@
 This comparison highlights relative risk (RR) as our primary performance indicator along with secondary measures of performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), as discussed more in Section~\ref{sec:metrics}.
+We observe improved performance over this benchmark across all metrics, and provide further comparisons between the OTS score and other models in Section~\ref{sec:model_comparison}.
+
 The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it uses time since the phase 2 transition as a predictive factor in addition to all others.
@@
 Utilizing our predictive models, we systematically identify target-disease pairs that have not yet been explored in clinical trials but exhibit a high likelihood of successful outcomes.
-This process involved ranking pairs based on their predicted success rates and filtering those with substantial supporting evidence but no prior clinical development attempts.
+This process involves ranking pairs based on their predicted success rates and filtering those with substantial supporting evidence but no prior clinical development attempts.
 
@@
 Figure~\ref{fig:opportunity_funnel} demonstrates this process visually, starting from the space of all present-day TD pairs.
-This funnel implies there are thousands of TD pairs that 1) have never been tested in clinical trials before, 2) have a 3x (RR=2.93) higher likelihood of advancing past phase 2 trials and 3) have a target that \textbf{has} been tested clinically (in at least phase 1 trials) for one or more of the
-modalities highlighted.
+This funnel implies there are thousands of TD pairs that 1) have never been tested in clinical trials before, 2) have a 3x (RR=2.93) higher likelihood of advancing past phase 2 trials and 3) have a target that \textbf{has} been tested clinically (in at least phase 1 trials) for one or more of the modalities highlighted.
 In other words, there are $\sim$2,400 small-molecule-enabled, $\sim$1,400 antibody-enabled, and 14 PROTAC-enabled TD pairs at the bottom of the funnel where the implied opportunity is greatest.

### 2024-06-06 paper/paper.tex
More review revisions

@@
   Daniel Himmelstein \hspace{0.2cm}
+  Daniel Frank \hspace{0.2cm} 
   \\ \\
@@
 \ul{To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates}.
-In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines.
+In contrast, we aim to establish a framework for identifying target-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines.
 This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.
@@
 Finally, we examine how many present-day target-disease pairs are likely to see success rates matching or exceeding those calibrated benchmarks, that are also undeveloped (i.e.
-have never been in clinical trials) and have a tractable target.
+have never been in clinical trials) and have a tractable target (i.e.
+with evidence suggesting that therapeutic modulation is possible).
 
@@
 The top margin shows the number of features in any one feature combination and the bottom margin shows the number of TD pairs that are a part of that same combination.
+There are \evaluationDatasetPairCount\ TD pairs in the evaluation dataset overall.
     (b) Same as (a) with proportional bar scaling that does not drop beyond a limit necessary to fit all margin labels and combinatorial groupings.
@@
 While deep learning methods for tabular data are competitive with gradient-boosted tree models \cite{mcelfresh2023neural, gorishniy2023revisiting, somepalli2021saint}, they are rarely superior without transfer learning \cite{mcelfresh2023neural, levin2023transfer, hollmann2023tabpfn}.
-They also have a number of operational drawbacks like necessitating GPUs and having less standard hyperparameter tuning procedures.
 Additionally, the data in this proof-of-concept is limited to Open Targets evidence scores that already reflect a large degree of feature engineering, thereby limiting the extent to which deep learning networks can identify more granular predictive signals.
@@
     (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and OTS (Open Targets composite scores). 
-    (b) Receiver operating characteristic (ROC) and average precision scores across the same 13 therapeutic areas with no limit on the number of rankings. 
+    (b) Receiver operating characteristic (ROC) and average precision scores across the same 13 therapeutic areas (each dot is one therapeutic area) with no limit on the number of rankings.
     See Supplementary~Figure~\ref{fig:relative_risk_by_ta} for raw data underlying (a).
@@
 We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement.
-See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model.
+See Section~\ref{sec:results_rankings} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model.
 This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.
@@
 \begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
+  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been under development after having reached phase 2.
   \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
@@
     Average precision (AP) and receiver operating characteristic (ROC) scores with relative risk (RR) at ranking cutoffs denoted by \colorbox{Gainsboro}{RR@N}.
+Dot sizes and colors (red=low score, green=high score) are both set according to the score shown in each cell.
     (a) Performance across constrained RDG models using different groups of features as described in Section~\ref{sec:models}
@@
 
-\subsection{Opportunities}
-\label{sec:results_opportunities}
+\subsection{Rankings}
+\label{sec:results_rankings}
 
@@
   \caption{
-    \textbf{Present-day target-disease pair opportunity funnel}.
+    \textbf{Present-day target-disease pair ranking funnel}.
     (a) All 2.53M TD pairs in Open Targets with evidence of some kind grouped by the maximum clinical trial phase reached for each as of the present day.
@@
 In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study.
-This is primarily because some evidence in Open Targets is not associated with publications.
+This is primarily because some evidence in Open Targets is not associated with publications, which prohibits us from associating it with a timestamp.
 Avoiding that limitation is appropriate in this context.
@@
 To address potential biases and inflation in our dataset, our methodology specifically considers the temporal emergence of evidence.
-By doing so, we ensure our predictions are grounded in data reflective of genuine biomedical insights, rather than artefacts of research focus or historical data accumulation trends.
+By doing so, we ensure our predictions are grounded in data reflective of genuine biomedical insights, rather than artifacts of research focus or historical data accumulation trends.
 
@@
 Pushing that even further to a 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively.
-These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} that there are more than 8,000 such opportunities (i.e.
+These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_rankings} that there are more than 8,000 such opportunities (i.e.
 untested, present-day TD pairs) where at least a 3x improvement in phase 2 success rates could be expected.
@@
   \item Not all Open Targets evidence can be associated with publication dates.
-This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_opportunities}.
+This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_rankings}.
+  \item The rankings of undeveloped TD pairs in Section~\ref{sec:results_rankings} are presented directly from Open Targets with no additional curation or vetting.
+We recognize that some of these TD pairs are for uninformative disease terms, e.g.
+the association between "CTLA4" and "autoimmune disease" in Supplementary~Figure~\ref{fig:top_opportunity_predictions}, and we opt not address these cases to best reflect the state of rankings within Open Targets.
+Several more computational steps are necessary in practice to identify a more enriched set of actionable TD pairs.
   \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}.
@@
   \item Estimating an absolute number of TD pairs associated with any criteria is difficult.
-In Section~\ref{sec:results_opportunities}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success.
+In Section~\ref{sec:results_rankings}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success.
 These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples.

### 2024-06-07 paper/paper.tex
More changes for reviews

@@
     (a) Presence and co-occurrence of select features for overlapping sets of TD pairs.
-The size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin).
+Each column of the visualization indicates a distinct grouping of TD pairs for which one combination of features is present.
+The rows denote features and the existence of any color bar other than white or grey indicates that the feature is present.
+For example, the first column from left-to-right shows that there are 4,295 TD pairs in the evaluation dataset with \textbf{only} target-specific and disease-specific clinical evidence (phase 2+ trials); i.e.
+no evidence exists that is specific to the TD pair for those cases.
+The size of each bar within a column is proportional to the number of TD pairs with that combination of features, and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin).
 The top margin shows the number of features in any one feature combination and the bottom margin shows the number of TD pairs that are a part of that same combination.
@@
 Features that are not specific to TD pairs are omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
+For more details on the origin and interpretation of these plots, see supervenn \cite{supervenn}.
   }
@@
 The use of such a metric is essential for properly assessing performance in this forecasting problem.
-While we also report more common measures of classifier performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), neither of these adequately capture behavior in the upper extremes of rankings due the sparsity with which TD pair evidence is present for pairs that have ever entered phase 2 trials.
+While we also report more common measures of classifier performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), neither of these adequately capture behavior in the upper extremes of rankings due to the sparsity with which TD pair evidence is present for pairs that have ever entered phase 2 trials.
 This sparsity is further exacerbated in this study by the constraint that most of that evidence must have existed \textit{before} such trials began.
@@
 Only the evaluation dataset is used to compute this average.
+The prefixes \colorbox{Gainsboro}{target}, \colorbox{Gainsboro}{disease} and \colorbox{Gainsboro}{target\_disease} are used to denote features that are specific to targets, diseases and target-disease pairs (respectively).
     (b) Feature coefficients, which are equivalent to the maximum effect that a feature can have due to all features being defined on the range [0, 1] or rescaled to it.
@@
     \put(100, 19) {(c)}
+    % Top section
+    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
+    \put(40, 57.5){\color{black}\line(0, -10){4}}
+    \put(40, 53.5){\color{black}\line(4, 0){4}}
     \put(50, 51){\color{black}\vector(0, -1){5}}
+    % Middle section
+    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
+    \put(33, 37.5){\color{black}\line(0, 10){4}}
+    \put(33, 41.5){\color{black}\line(3, 0){3}}
     \put(74, 39){\color{black}\vector(0, -1){4.5}}
-    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
-    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
+    % Bottom section
     \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x+ more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
+    \put(82, 23){\color{black}\line(5, 0){5}}
+    \put(87, 23){\color{black}\line(0, -21){21}}
+    \put(85, 2){\color{black}\line(2, 0){2}}
+    % \put(33, 41.5){\color{black}\line(3, 0){3}}
   \end{overpic}
@@
 Examples of antibody-enabled TD pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.
+Note that these are intentionally not curated or filtered in order to capture rankings as they would be seen for the corresponding version of Open Targets (23.12).
+This is discussed more in Section~\ref{sec:limitations}.
 

### 2024-07-31 paper/paper.tex
Update discussion and add data availability statement

@@
   Daniel Frank \hspace{0.2cm} 
+  Nick Miller \hspace{0.2cm} 
   \\ \\
@@
 
+% The BIO report estimate used here comes from Figure 2, which says 4933 programs
+% entered phase 2 across all indications in the 2011–2020 timeframe.
 An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery.
@@
 untested, present-day TD pairs) where at least a 3x improvement in phase 2 success rates could be expected.
+This suggests that a substantial fraction of the 493 drug programs entering phase 2 trials each year \cite{BIOReport} could potentially be replaced by programs from this pool.
 
-Another way to quantify the impact that employing superior target prioritization strategies could have, is to consider what would happen if this was done across the entire pharmaceutical development industry.
-Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year.
-This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ...
-(TODO: finish with Adam)
-
-Throughout the majority of this study, we focus primarily on performance among the top \topRankingPct\% of TD pairs.
+Throughout the majority of this study, we focus on performance among the top \topRankingPct\% of TD pairs.
 Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}.
@@
 Our primary results are based on the 23.12 Open Targets release, but we also use releases 23.09 and 23.06 as part of our sensitivity analysis in section \ref{sec:sensitivity}.
+All code used is available at \url{https://github.com/related-sciences/clinical_advancement_paper}.
 
@@
 \end{enumerate}
+
+\section{Data availability}
+
+All code, data and analysis used for this study, when not noted as proprietary, is available at \url{https://github.com/related-sciences/clinical_advancement_paper}.
 
